---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, comment = "#>")
```

# bigQTL

Scalable conditional cis-QTL analysis on file-backed matrices (FBM) from the
bigstatsr/bigsnpr ecosystem: stepwise conditioning, all-but-one conditioning, and
eigenMT multiple-testing correction, with results written as partitioned Apache
Parquet datasets.

Repository: <https://github.com/ImmuneAxisa/bigQTL>

## Installation

```{r}
remotes::install_github("ImmuneAxisa/bigQTL")
```

All runtime dependencies are declared in `DESCRIPTION` and installed automatically.

## Features

- **Conditional QTL analysis** — `bigQTL()` (all-in-one, with automatic PC
  computation), `run_conditional_qtl()` (stepwise + all-but-one conditioning),
  `marginalQTL()` (marginal-only scan)
- **eigenMT multiple-testing correction** — `eigenMT_batch()`, `eigenMT_gene()`,
  `eigenMT_correct()`, with a dependency-free Ledoit-Wolf shrinkage estimator
- **Data preparation** — `compute_geno_pcs()`, `compute_pheno_pcs()`, `rint()`
- **`bigPheno`** — an FBM-backed S3 class for phenotype/expression matrices, mirroring
  bigsnpr's `bigSNP`

## Quickstart

```{r}
library(bigQTL)

bigpheno <- bigPheno(expr_mat)  # samples x phenotypes

res <- bigQTL(
  bigpheno    = bigpheno,
  bigsnp      = bigsnp,          # a bigsnpr::bigSNP object
  pheno_coord = pheno_coord,     # data.frame(pheno_name, chromosome, start, end)
  design_base = design_matrix,   # covariates; rownames = sample IDs
  output_dir  = "output/"
)

res$stepwise    # Arrow dataset: output/stepwise/pheno=<name>/part-0.parquet
res$allbutone   # Arrow dataset: output/allbutone/pheno=<name>/part-0.parquet
```

See `vignette("bigQTL-intro")` for a complete walkthrough (simulated data, stepwise +
all-but-one inspection, eigenMT correction), and `vignettes/articles/
eigenMT_validation.Rmd` for validation against the original Python eigenMT.

## License & citation

CC BY 4.0 — see `LICENSE.md`. If you use `bigQTL` in a publication, please cite the
repository and the underlying methods (eigenMT: Davis et al. 2016; Ledoit-Wolf
shrinkage: Ledoit & Wolf 2004).

## Contact

Issues and questions: <https://github.com/ImmuneAxisa/bigQTL/issues>
