---
title: "Getting Started with bigQTL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bigQTL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette walks through the main capabilities of `bigQTL` on a small simulated
dataset: constructing the package's FBM-backed data objects, running conditional
cis-QTL mapping (stepwise conditioning and all-but-one), and applying eigenMT
multiple-testing correction. For the full architecture and function reference see
`dev/ARCHITECTURE.md` in the package repository.

```{r setup}
library(bigQTL)
```

## 1. Simulate genotype and phenotype data

`bigQTL` operates on file-backed matrices (FBM) from the bigstatsr/bigsnpr ecosystem:
a `bigSNP` object (genotypes) and a `bigPheno` object (phenotypes), matched by sample
ID. Here we simulate 100 samples, 300 SNPs on one chromosome, and a phenotype panel
with two genes of interest carrying a real, independent two-SNP signal (`geneA`
depends on both `snp_150` and `snp_50`), plus `geneB`/`geneC` as unassociated
controls. We also add 150 unrelated background phenotypes — in a real analysis these
would be the rest of the expression panel; including them here gives
`compute_pheno_pcs()` something realistic to compute PCs over, rather than PCA-ing
just the 3 genes we're about to test (which would just rediscover each gene's own
signal).

```{r simulate}
set.seed(42)

n_samples <- 100
n_snps <- 300
n_bg_phenos <- 150

# bigsnpr::snp_fake() allocates a properly-typed bigSNP object (FBM.code256
# genotypes); we fill in the genotype values ourselves.
bigsnp <- bigsnpr::snp_fake(n_samples, n_snps)
bigsnp$genotypes[] <- sample(0:2, n_samples * n_snps, replace = TRUE,
                              prob = c(0.25, 0.5, 0.25))
bigsnp$map$chromosome   <- as.character(bigsnp$map$chromosome)
bigsnp$map$physical.pos <- seq(1e6, 2e6, length.out = n_snps)
sample_ids <- bigsnp$fam$sample.ID

# Two independent causal SNPs for geneA
g1 <- bigsnp$genotypes[, 150]  # "snp_150"
g2 <- bigsnp$genotypes[, 50]   # "snp_50"

bg_mat <- matrix(rnorm(n_samples * n_bg_phenos), n_samples, n_bg_phenos)
colnames(bg_mat) <- paste0("bg", seq_len(n_bg_phenos))

expr_mat <- matrix(rnorm(n_samples * 3), n_samples, 3)
expr_mat[, 1] <- expr_mat[, 1] + 2.0 * g1 + 1.6 * g2
colnames(expr_mat) <- c("geneA", "geneB", "geneC")

full_mat <- cbind(expr_mat, bg_mat)
rownames(full_mat) <- sample_ids

bigpheno <- bigPheno(full_mat)
```

`pheno_coord` gives the genomic coordinates used to define each phenotype's cis
window; only phenotypes listed here are tested (the background phenotypes are used
for PC computation only, via `bigQTL()`, but not mapped):

```{r pheno-coord}
pheno_coord <- data.frame(
  pheno_name = c("geneA", "geneB", "geneC"),
  chromosome = "1",
  start      = c(1.5e6, 1.2e6, 1.8e6),
  end        = c(1.5e6, 1.2e6, 1.8e6)
)

# design_base: covariates, row names = sample IDs (required)
design_base <- data.frame(
  age = rnorm(n_samples, mean = 50, sd = 10),
  row.names = sample_ids
)
```

## 2. Marginal QTL scan

`marginalQTL()` tests every cis-SNP against every phenotype with no conditioning —
useful as a quick first look:

```{r marginal}
out_dir_marginal <- tempfile("marginal_")
res_marginal <- marginalQTL(
  bigpheno    = bigpheno,
  bigsnp      = bigsnp,
  pheno_coord = pheno_coord,
  design_base = design_base,
  min_snps    = 10,
  output_dir  = out_dir_marginal
)

marginal_df <- dplyr::collect(res_marginal$stepwise)
head(marginal_df[order(marginal_df$pvalue), c("pheno", "snp", "beta", "pvalue", "fdr")])
```

## 3. Full conditional analysis with `bigQTL()`

`bigQTL()` is the all-in-one wrapper: it computes phenotype PCs (from the full
phenotype panel) and genotype PCs, appends them to `design_base` as covariates, then
runs `run_conditional_qtl()` — stepwise conditioning followed by all-but-one, for
every phenotype in `pheno_coord`.

```{r bigqtl, message=FALSE}
out_dir <- tempfile("bigqtl_")
res <- bigQTL(
  bigpheno       = bigpheno,
  bigsnp         = bigsnp,
  pheno_coord    = pheno_coord,
  design_base    = design_base,
  n_pheno_pcs    = 3,
  n_geno_pcs     = 2,
  pval_threshold = 1e-4,
  min_snps       = 10,
  output_dir     = out_dir
)
```

`res$stepwise` and `res$allbutone` are Arrow datasets (partitioned Parquet under
`out_dir`), queryable lazily with `dplyr`/`arrow` before pulling results into memory:

```{r inspect-stepwise}
stepwise_df <- res$stepwise |>
  dplyr::filter(pheno == "geneA") |>
  dplyr::select(step, conditioning_snps, snp, beta, pvalue) |>
  dplyr::collect()

# The two real signals, snp_150 and snp_50, are recovered as independent hits;
# stepwise conditioning stops as soon as no further SNP clears pval_threshold
# (see dev/CODE_REVIEW.md for the history of a prototype bug where one extra,
# non-significant step used to leak into results here).
stepwise_df[stepwise_df$pvalue < 1e-3, ]
```

With two independent signals for `geneA`, all-but-one conditioning has something to
do — for each signal, it tests that SNP once more while conditioning on every *other*
independent signal, to check it holds up on its own:

```{r inspect-allbutone}
allbutone_df <- res$allbutone |>
  dplyr::filter(pheno == "geneA") |>
  dplyr::collect()

unique(allbutone_df[, c("indep", "conditioning_snps")])
```

## 4. eigenMT multiple-testing correction

Within a cis window, nearby SNPs are correlated (LD), so treating every SNP as an
independent test is overly conservative. `eigenMT_batch()` estimates the effective
number of independent tests (`m_eff`) per phenotype from the local LD structure, and
`eigenMT_correct()` applies a Bonferroni-style correction using it instead of the raw
SNP count:

```{r eigenmt}
m_eff <- eigenMT_batch(
  bigsnp      = bigsnp,
  pheno_coord = pheno_coord,
  ind.row     = seq_len(n_samples)
)
m_eff

marginal_df$m_eff       <- m_eff$m_eff[match(marginal_df$pheno, m_eff$pheno_name)]
marginal_df$pval_eigenmt <- eigenMT_correct(marginal_df$pvalue, marginal_df$m_eff)

head(marginal_df[order(marginal_df$pvalue),
                  c("pheno", "snp", "pvalue", "m_eff", "pval_eigenmt")])
```

## Next steps

- `vignettes/articles/eigenMT_validation.Rmd` (built as a pkgdown article, not part of
  the installed package) validates the `eigenMT` implementation against the original
  Python version on real data.
- `dev/ARCHITECTURE.md` documents the full function catalog, data model, and
  parallelism model (`ncores` vs `ncores_phenos`).
