---
title: "eigenMT Validation: R vs Python"
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Overview

This vignette validates the R implementation of eigenMT (implemented in
`bigQTL`) against the original Python eigenMT
([joed3/eigenMT](https://github.com/joed3/eigenMT)) by running both on
identical simulated genotype data with realistic LD structure and comparing
M\_eff per gene.

All code chunks use `eval = FALSE` and are intended to be run interactively.

## Prerequisites

You need a conda environment named `"eigenMT"` with the required Python
packages. Create it once with:

```{r conda-setup}
# if you do not have a conda install you can run reticulate::install_miniconda()
library(reticulate)
conda_create("eigenMT",
             c("numpy", "scipy", "Scikit-learn", "pandas"),
             python_version = "3.8")
```

Then load the package:

```{r load-pkg}
library(bigQTL)
library(bigstatsr)
library(bigsnpr)
```

# Simulation Parameters

```{r params}
set.seed(2000)

N_SAMPLES  <- 500L   # number of individuals
N_SNPS     <- 800L   # total SNPs on chr1
BLOCK_SIZE <- 20L    # SNPs per LD block
FLIP_PROB  <- 0.10   # probability of flipping an allele within a block
VAR_THRESH <- 0.99
EIGEN_WIN  <- 200L
CIS_DIST   <- 1e6
```

# 1. Simulate Genotype Data

```{r simulate-geno}
snp_pos <- round(seq(1e6, 2e6, length.out = N_SNPS))
snp_ids <- paste0("rs", seq_len(N_SNPS))
chr_vec <- rep("1", N_SNPS)

maf_vec  <- stats::runif(N_SNPS, 0.05, 0.45)
geno_mat <- matrix(NA_integer_, nrow = N_SAMPLES, ncol = N_SNPS)

n_blocks <- ceiling(N_SNPS / BLOCK_SIZE)
for (b in seq_len(n_blocks)) {
  block_start <- (b - 1L) * BLOCK_SIZE + 1L
  block_end   <- min(b * BLOCK_SIZE, N_SNPS)
  block_snps  <- block_start:block_end

  founder_maf  <- maf_vec[block_start]
  founder_geno <- stats::rbinom(N_SAMPLES, 2L, founder_maf)
  geno_mat[, block_start] <- founder_geno

  for (j in block_snps[-1]) {
    flipped  <- stats::rbinom(N_SAMPLES, 1L, FLIP_PROB)
    fresh    <- stats::rbinom(N_SAMPLES, 2L, maf_vec[j])
    new_geno <- ifelse(flipped == 1L, fresh, founder_geno)
    geno_mat[, j] <- new_geno
  }
}

sample_ids <- paste0("sample_", seq_len(N_SAMPLES))

message(sprintf("Simulated %d x %d genotype matrix (%d LD blocks of ~%d SNPs)",
                N_SAMPLES, N_SNPS, n_blocks, BLOCK_SIZE))
```

# 2. Define Gene Positions

```{r gene-positions}
features_coord <- data.frame(
  feature_name = c("GeneA", "GeneB", "GeneC"),
  chromosome   = c("1",     "1",     "1"),
  start        = c(1.10e6,  1.45e6,  1.80e6),
  end          = c(1.12e6,  1.47e6,  1.82e6),
  stringsAsFactors = FALSE
)
```

# 3. Write Python eigenMT Input Files

```{r write-python-inputs}
tmp_dir <- file.path(tempdir(), "eigenMT_validation")
dir.create(tmp_dir, recursive = TRUE, showWarnings = FALSE)

## Genotypes: SNPs x samples
geno_for_py <- cbind(snp_ids, as.data.frame(t(geno_mat)))
colnames(geno_for_py) <- c("id", sample_ids)
geno_file <- file.path(tmp_dir, "genotypes.txt")
write.table(geno_for_py, geno_file, sep = "\t", quote = FALSE,
            row.names = FALSE, col.names = TRUE)

## SNP positions
gen_pos <- data.frame(snp = snp_ids, chr = chr_vec, pos = snp_pos,
                      stringsAsFactors = FALSE)
gen_pos_file <- file.path(tmp_dir, "gen.positions.txt")
write.table(gen_pos, gen_pos_file, sep = "\t", quote = FALSE,
            row.names = FALSE, col.names = TRUE)

## Gene positions
phe_pos <- data.frame(
  geneid = features_coord$feature_name,
  chr    = features_coord$chromosome,
  s1     = features_coord$start,
  s2     = features_coord$end,
  stringsAsFactors = FALSE
)
phe_pos_file <- file.path(tmp_dir, "phe.positions.txt")
write.table(phe_pos, phe_pos_file, sep = "\t", quote = FALSE,
            row.names = FALSE, col.names = TRUE)

## Fake cis eQTL results (one row per cis SNP-gene pair)
eqtl_rows <- list()
for (i in seq_len(nrow(features_coord))) {
  gene_row  <- features_coord[i, ]
  win_start <- gene_row$start - CIS_DIST
  win_end   <- gene_row$end   + CIS_DIST
  cis_mask  <- (chr_vec == gene_row$chromosome) &
               (snp_pos >= win_start) & (snp_pos <= win_end)
  cis_snps  <- snp_ids[cis_mask]
  if (length(cis_snps) > 0) {
    np <- length(cis_snps)
    eqtl_rows[[i]] <- data.frame(
      SNP       = cis_snps,
      gene      = rep(gene_row$feature_name, np),
      beta      = stats::rnorm(np),
      `t-stat`  = stats::rnorm(np),
      `p-value` = stats::runif(np),
      FDR       = stats::runif(np),
      stringsAsFactors = FALSE,
      check.names = FALSE
    )
  }
}
eqtl_df   <- do.call(rbind, eqtl_rows)
eqtl_file <- file.path(tmp_dir, "cis.eqtls.txt")
write.table(eqtl_df, eqtl_file, sep = "\t", quote = FALSE,
            row.names = FALSE, col.names = TRUE)
```

# 4. Build bigSNP Object

```{r build-bigsnp}
fbm_file <- file.path(tmp_dir, "geno.bk")
geno_fbm <- bigstatsr::FBM(
  nrow = N_SAMPLES, ncol = N_SNPS, type = "integer",
  backingfile = sub("\\.bk$", "", fbm_file),
  create_bk   = TRUE
)
geno_fbm[] <- geno_mat

snp_map <- data.frame(
  chromosome   = chr_vec,
  marker.ID    = snp_ids,
  genetic.dist = rep(0, N_SNPS),
  physical.pos = snp_pos,
  allele1      = rep("A", N_SNPS),
  allele2      = rep("C", N_SNPS),
  stringsAsFactors = FALSE
)

snp_fam <- data.frame(
  family.ID   = sample_ids,
  sample.ID   = sample_ids,
  paternal.ID = rep(0, N_SAMPLES),
  maternal.ID = rep(0, N_SAMPLES),
  sex         = rep(0, N_SAMPLES),
  affection   = rep(0, N_SAMPLES),
  stringsAsFactors = FALSE
)

bigsnp_obj <- structure(
  list(genotypes = geno_fbm, map = snp_map, fam = snp_fam),
  class = "bigSNP"
)

ind.row <- seq_len(N_SAMPLES)
```

# 5. Embed and Run Python eigenMT

The eigenMT Python source is embedded below and written to a temp file,
then executed via `reticulate::conda_run2()` using the `"eigenMT"` conda
environment.

```{r run-python}
python_script_file <- file.path(tmp_dir, "eigenMT.py")
download.file("https://raw.githubusercontent.com/joed3/eigenMT/refs/heads/master/eigenMT.py",
              python_script_file)

py_out_file <- file.path(tmp_dir, "eigenMT_output.txt")

reticulate::conda_run2(
  "python",
  args = c(
    python_script_file,
    "--CHROM",      "1",
    "--QTL",        eqtl_file,
    "--GEN",        geno_file,
    "--GENPOS",     gen_pos_file,
    "--PHEPOS",     phe_pos_file,
    "--OUT",        py_out_file,
    "--cis_dist",   format(CIS_DIST, scientific = FALSE),
    "--var_thresh", format(VAR_THRESH, scientific = FALSE),
    "--window",     format(EIGEN_WIN, scientific = FALSE)
  ),
  envname = "eigenMT"
)

python_results <- read.table(py_out_file, header = TRUE,
                             sep = "\t", stringsAsFactors = FALSE)
message(sprintf("Python eigenMT: %d genes", nrow(python_results)))
```

# 6. Run R eigenMT

```{r run-r-eigenmt}
r_results <- lapply(seq_len(nrow(features_coord)), function(i) {
  gene_row <- features_coord[i, ]
  cis_result <- get_cis_snps(
    bigsnp     = bigsnp_obj,
    gene_chr   = gene_row$chromosome,
    gene_start = gene_row$start,
    gene_end   = gene_row$end,
    cis_window = CIS_DIST
  )
  m_eff <- eigenMT_gene(
    bigsnp      = bigsnp_obj,
    snp_indices = cis_result$indices,
    ind.row     = ind.row,
    var_thresh  = VAR_THRESH,
    window      = EIGEN_WIN
  )
  list(gene = gene_row$feature_name,
       n_cis_snps = length(cis_result$indices),
       m_eff = m_eff)
})

r_df <- do.call(rbind, lapply(r_results, as.data.frame))
print(r_df)
```

# 7. Unit Tests

```{r unit-tests}
n_failures <- 0L
use_condaenv("eigenMT")
eigenMTpy <- reticulate::import_from_path("eigenMT.py")
## 7a. count_eigenvalues vs Python find_num_eigs
py_find_num_eigs <- function(eigenvalues, variance, var_thresh) {
  eigenvalues <- sort(eigenvalues, decreasing = TRUE)
  running_sum <- 0
  counter <- 0L
  while (running_sum < variance * var_thresh) {
    running_sum <- running_sum + eigenvalues[counter + 1L]
    counter <- counter + 1L
  }
  counter
}
test_eigs <- c(6.5, 2.3, 1.1, 1, 0.8, 0.6, 0.5, 0.4, 0.3, 0.2, 0.11, 0.1, 0.05, 0.04)
r_count  <- count_eigenvalues(test_eigs, 0.99)
py_count <- py_find_num_eigs(test_eigs, length(test_eigs), 0.99)
stopifnot("count_eigenvalues matches find_num_eigs" = r_count == py_count)
message(sprintf("  count_eigenvalues: PASS (both return %d)", r_count))

## 7b. big_cor vs base cor()
test_cols   <- 1:30
base_cor    <- cor(geno_mat[, test_cols])
fbm_cor     <- bigstatsr::big_cor(geno_fbm, ind.row = ind.row,
                                  ind.col = test_cols)[]
frob_diff   <- sqrt(sum((base_cor - fbm_cor)^2))
stopifnot("big_cor matches cor()" = frob_diff < 1e-8)
message(sprintf("  big_cor vs cor(): PASS (Frob diff = %.2e)", frob_diff))

## 7c. lw_shrink_cor sanity checks
R_test    <- cor(geno_mat[, 1:20])
shrunk    <- lw_shrink_cor(R_test, N_SAMPLES)
eigs      <- eigen(shrunk, symmetric = TRUE, only.values = TRUE)$values
stopifnot("shrunk matrix is PSD"        = all(eigs >= -1e-10))
stopifnot("shrunk matrix diagonal is 1" = all(abs(diag(shrunk) - 1) <= 1e-10))
message("  lw_shrink_cor: PASS (PSD, diagonal = 1)")

## 7d. eigenMT_gene edge cases
m1 <- eigenMT_gene(bigsnp_obj, snp_indices = 1L,          ind.row = ind.row)
m0 <- eigenMT_gene(bigsnp_obj, snp_indices = integer(0),  ind.row = ind.row)
stopifnot("single-SNP M_eff = 1" = m1 == 1L)
stopifnot("zero-SNP M_eff = 0"   = m0 == 0L)
message("  eigenMT_gene edge cases: PASS")

message(sprintf("\nUnit tests: %d failures", n_failures))
```

# 8. Compare R vs Python M\_eff

```{r compare}
MEFF_TOL <- 0.20  # 20% relative difference tolerance

comparison <- do.call(rbind, lapply(features_coord$feature_name, function(gene) {
  r_row  <- r_df[r_df$gene == gene, ]
  py_row <- python_results[python_results$gene == gene, ]
  if (nrow(r_row) == 0 || nrow(py_row) == 0) return(NULL)
  m_r  <- r_row$m_eff[1]
  m_py <- py_row$TESTS[1]
  data.frame(gene = gene,
             n_cis_snps = r_row$n_cis_snps[1],
             m_eff_r  = m_r,
             m_eff_py = m_py,
             rel_diff = abs(m_r - m_py) / max(m_py, 1),
             stringsAsFactors = FALSE)
}))

print(comparison)

bad <- comparison$rel_diff > MEFF_TOL
if (any(bad)) {
  stop(sprintf("M_eff differs by >%d%% for: %s",
               as.integer(MEFF_TOL * 100),
               paste(comparison$gene[bad], collapse = ", ")))
}
message(sprintf("All %d genes pass M_eff tolerance (<= %d%% relative diff)",
                nrow(comparison), as.integer(MEFF_TOL * 100)))
```

# 9. Clean Up

```{r cleanup}
unlink(tmp_dir, recursive = TRUE)
message("Temp files removed.")
```
