---
title: "eigenMT Validation: R vs Python"
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
```

# Overview

This vignette validates the R implementation of eigenMT (implemented in
`bigQTL`) against the **original** Python eigenMT
([joed3/eigenMT](https://github.com/joed3/eigenMT)) by running both on the
real example dataset shipped with that repository and diagnosing the sources
of any difference.

All code chunks use `eval = FALSE` and are intended to be run interactively.

## Known implementation differences (updated after audit)

| # | Area | Python | R `bigQTL` (current) | Severity |
|---|------|--------|-----------------------|----------|
| 1 | **Shrinkage estimator — input** | `LedoitWolf.fit()` on raw centred genotype matrix (n × p); produces shrunk **covariance**, then normalised to correlation | OAS formula applied to pre-computed **correlation** matrix — OAS on a correlation matrix yields near-zero alpha, so almost no shrinkage occurs | 🔴 **Critical — root cause of the observed 100× difference** |
| 2 | **Shrinkage formula** | Ledoit-Wolf 2004 analytic formula | OAS (Chen et al. 2010) — a different estimator even if applied to the same input | 🔴 Major |
| 3 | **Missing genotype imputation** | `NA` → column mean before any computation | No imputation; `bigstatsr::big_cor()` propagates `NA` | 🟠 Moderate |
| 4 | **SNP ordering** | By genomic position | By FBM column index | 🟡 Minor (equivalent when FBM is position-sorted) |
| 5 | `count_eigenvalues` normalisation | Window size `m` as `variance` | `sum(eigenvalues)`; equivalent since `trace(cor) == p` | 🟢 Equivalent |
| 6 | Window traversal | Disjoint, `break` on 1-SNP tail | Same | 🟢 Equivalent |

**Key insight on issue #1:** The OAS formula was derived for covariance
matrices. When applied directly to a correlation matrix (diagonal = 1 by
construction, so `tr(R)^2/p = p` is fixed regardless of off-diagonal
structure), the formula systematically underestimates the shrinkage intensity.
For a 200-SNP window with n ≈ 200 samples, OAS-on-correlation gives alpha ≈ 0,
meaning the "shrunk" matrix is almost identical to the raw correlation matrix.
This makes almost all eigenvalues needed to reach the 99% threshold →
M_eff ≈ n_cis_snps instead of ~40.

## Fix: use `cvCovEst::linearShrinkLWEst` per window

The correct approach is to:
1. Extract the raw genotype submatrix (n × p) for each window
2. Apply `cvCovEst::linearShrinkLWEst()` which implements the LW 2004
   formula on the sample covariance
3. Normalise to a correlation matrix: `D^{-1/2} Σ_shrunk D^{-1/2}`
4. Eigendecompose and count

Because we use per-window submatrices (at most 200 SNPs × n samples), memory
usage is bounded and does not require materialising the full genotype matrix.

## Prerequisites

```{r conda-setup}
library(reticulate)


if (!any("eigenMT" %in% conda_list()[["name"]])) {
  
  conda_create(
    "eigenMT",
    packages = c("numpy", "scipy", "scikit-learn<1.2", "pandas"),
    python_version = "3.8"
  )
  
}

```

```{r load-pkg}
library(bigQTL)
library(bigstatsr)
library(bigsnpr)
# install.packages("cvCovEst")  # if not already installed
library(cvCovEst)
```

# 1. Download and Decompress the Example Data

```{r download-example}
tmp_dir <- file.path(tempdir(), "eigenMT_validation")
dir.create(tmp_dir, recursive = TRUE, showWarnings = FALSE)

tgz_url  <- paste0(
  "https://github.com/joed3/eigenMT/raw/",
  "f91e2837d34403cff48c9a248a7d4fb3fac98541/example.tar.gz"
)
tgz_file <- file.path(tmp_dir, "example.tar.gz")
utils::download.file(tgz_url, destfile = tgz_file, mode = "wb", quiet = FALSE)
message(sprintf("Downloaded %.2f MB", file.info(tgz_file)$size / 1e6))

utils::untar(tgz_file, exdir = tmp_dir)
example_dir <- file.path(tmp_dir, "example")
stopifnot(dir.exists(example_dir))
message("Example files: ", paste(list.files(example_dir), collapse = ", "))
```

# 2. Download the Original Python Script

```{r download-python}
py_script_url <- paste0(
  "https://raw.githubusercontent.com/joed3/eigenMT/",
  "f91e2837d34403cff48c9a248a7d4fb3fac98541/eigenMT.py"
)
py_script_file <- file.path(tmp_dir, "eigenMT.py")
utils::download.file(py_script_url, destfile = py_script_file, quiet = FALSE)
message("Downloaded eigenMT.py")
```

# 3. Read the Example Data

```{r read-data}
geno_raw <- read.table(file.path(example_dir, "genotypes.txt"),
                       header = TRUE, sep = "\t",
                       check.names = FALSE, stringsAsFactors = FALSE)
message(sprintf("Genotype matrix: %d SNPs x %d samples",
                nrow(geno_raw), ncol(geno_raw) - 1L))

gen_pos <- read.table(file.path(example_dir, "gen.positions.txt"),
                      header = TRUE, sep = "\t", stringsAsFactors = FALSE)
phe_pos <- read.table(file.path(example_dir, "phe.positions.txt"),
                      header = TRUE, sep = "\t", stringsAsFactors = FALSE)
eqtl    <- read.table(file.path(example_dir, "cis.eqtls.txt"),
                      header = TRUE, sep = "\t", stringsAsFactors = FALSE)

message(sprintf("SNPs: %d | Genes: %d | eQTL rows: %d",
                nrow(gen_pos), nrow(phe_pos), nrow(eqtl)))
message(sprintf("Chromosomes in data: %s",
                paste(unique(gen_pos[[2L]]), collapse = ", ")))
message(sprintf("eQTL file columns: %s",
                paste(colnames(eqtl), collapse = ", ")))

table(eqtl$gene)
```

# 4. Determine chromosome

The example data may be from chr19. Detect the chromosome automatically.

```{r detect-chrom}
CHROM <- as.character(unique(gen_pos[[2L]])[1L])
message(sprintf("Using chromosome: %s", CHROM))
```

# 5. Run Python eigenMT

```{r run-python}
py_out_file <- file.path(tmp_dir, "eigenMT_output_py.txt")

reticulate::conda_run2(
  "python",
  args = c(
    py_script_file,
    "--CHROM",      CHROM,
    "--QTL",        file.path(example_dir, "cis.eqtls.txt"),
    "--GEN",        file.path(example_dir, "genotypes.txt"),
    "--GENPOS",     file.path(example_dir, "gen.positions.txt"),
    "--PHEPOS",     file.path(example_dir, "phe.positions.txt"),
    "--OUT",        py_out_file,
    "--cis_dist",   "1000000",
    "--var_thresh", "0.99",
    "--window",     "200"
  ),
  envname = "eigenMT"
)

python_results <- read.table(py_out_file, header = TRUE,
                             sep = "\t", stringsAsFactors = FALSE)
message(sprintf("Python eigenMT: %d genes", nrow(python_results)))
print(head(python_results))
```

# 6. Build bigSNP Object

```{r build-bigsnp}
sample_ids  <- colnames(geno_raw)[-1L]
N_SAMPLES   <- length(sample_ids)
snp_ids     <- as.character(geno_raw[[1L]])
geno_mat    <- as.matrix(geno_raw[, -1L, drop = FALSE])
storage.mode(geno_mat) <- "double"   # keep as double for LW estimator

## Samples x SNPs
geno_mat_t  <- t(geno_mat)
N_SNPS      <- ncol(geno_mat_t)

## SNP position lookup
snp_order <- match(snp_ids, as.character(gen_pos[[1L]]))
if (any(is.na(snp_order)))
  warning(sum(is.na(snp_order)), " SNP IDs not found in gen.positions.txt")
chr_vec <- as.character(gen_pos[[2L]][snp_order])
pos_vec <- as.numeric(gen_pos[[3L]][snp_order])

## Impute NA -> column mean (mirrors Python)
for (j in seq_len(N_SNPS)) {
  col_j <- geno_mat_t[, j]
  na_m  <- is.na(col_j)
  if (any(na_m))
    geno_mat_t[na_m, j] <- mean(col_j, na.rm = TRUE)
}

## FBM
fbm_file <- file.path(tmp_dir, "geno")
geno_fbm <- bigstatsr::FBM(
  nrow = N_SAMPLES, ncol = N_SNPS, type = "double",
  backingfile = fbm_file, create_bk = TRUE
)
geno_fbm[] <- geno_mat_t

snp_map <- data.frame(
  chromosome   = chr_vec,  marker.ID    = snp_ids,
  genetic.dist = 0,        physical.pos = pos_vec,
  allele1      = "A",      allele2      = "C",
  stringsAsFactors = FALSE
)
snp_fam <- data.frame(
  family.ID = sample_ids, sample.ID   = sample_ids,
  paternal.ID = 0L,       maternal.ID = 0L,
  sex = 0L,               affection   = 0L,
  stringsAsFactors = FALSE
)
bigsnp_obj <- structure(list(genotypes = geno_fbm, map = snp_map, fam = snp_fam),
                        class = "bigSNP")
ind.row <- seq_len(N_SAMPLES)
message(sprintf("bigSNP: %d samples x %d SNPs", N_SAMPLES, N_SNPS))
```

# 7. Gene Coordinates

```{r build-features}
genes_in_eqtl  <- unique(eqtl[[2L]])
features_coord <- data.frame(
  feature_name = as.character(phe_pos[[1L]]),
  chromosome   = as.character(phe_pos[[2L]]),
  start        = as.numeric(phe_pos[[3L]]),
  end          = as.numeric(phe_pos[[4L]]),
  stringsAsFactors = FALSE
)
features_coord <- features_coord[
  features_coord$chromosome == CHROM &
    features_coord$feature_name %in% genes_in_eqtl, ]
message(sprintf("Genes to process: %d", nrow(features_coord)))
```

# 8. Shrinkage Diagnostic — Side-by-Side Comparison on One Window

Before running the full analysis, inspect the shrinkage behaviour of the
three approaches on a small window of real data so we can understand why
OAS-on-correlation fails.

```{r shrinkage-diagnostic}
## Take the first 50 SNPs as a test window
diag_cols <- seq_len(min(50L, N_SNPS))
win_geno  <- geno_mat_t[, diag_cols]   # n x p (samples x SNPs)

## --- Method A: OAS on correlation (current bigQTL) ---
R_raw <- cor(win_geno)
p     <- ncol(win_geno)
n     <- nrow(win_geno)
trR2  <- sum(R_raw^2)
num   <- (1 - 2/p) * trR2 + p^2
den   <- (n + 1 - 2/p) * (trR2 - p)
alpha_oas <- max(0, min(1, num / den))
message(sprintf("OAS-on-correlation alpha: %.6f  (near 0 = almost no shrinkage)", alpha_oas))

## --- Method B: LW 2004 on raw data (Python-faithful) ---
lw_cov   <- cvCovEst::linearShrinkLWEst(dat = win_geno)
## Convert shrunk covariance to correlation
d_inv    <- diag(1 / sqrt(diag(lw_cov)))
lw_cor   <- d_inv %*% lw_cov %*% d_inv
## The LW shrinkage intensity (b_n2/d_n2) can be read from the off-diagonal shrinkage:
## alpha_lw = 1 - (off-diag of lw_cov) / (off-diag of sample_cov)
sample_cov  <- cov(win_geno)
alpha_lw_approx <- 1 - mean(lw_cov[row(lw_cov) != col(lw_cov)] /
                             sample_cov[row(sample_cov) != col(sample_cov)])
message(sprintf("LW2004-on-raw-data effective shrinkage: ~%.6f", alpha_lw_approx))

## --- Compare eigenvalue distributions ---
eigs_unshrunk <- eigen(R_raw,         symmetric=TRUE, only.values=TRUE)$values
eigs_oas      <- eigen(lw_shrink_cor(R_raw, n), symmetric=TRUE, only.values=TRUE)$values
eigs_lw       <- eigen(lw_cor,        symmetric=TRUE, only.values=TRUE)$values
eigs_unshrunk[eigs_unshrunk < 0] <- 0
eigs_oas[eigs_oas < 0]           <- 0
eigs_lw[eigs_lw < 0]             <- 0

m_eff_unshrunk <- count_eigenvalues(eigs_unshrunk, 0.99)
m_eff_oas      <- count_eigenvalues(eigs_oas,      0.99)
m_eff_lw       <- count_eigenvalues(eigs_lw,       0.99)

message(sprintf(
  "M_eff (p=%d window) — unshrunk: %d | OAS-on-cor: %d | LW-on-raw: %d",
  p, m_eff_unshrunk, m_eff_oas, m_eff_lw
))
```

# 9. Python-Faithful R Eigenvalue Counter

A standalone helper that replicates exactly what Python does inside
`bf_eigen_windows` for a single window: takes the raw genotype submatrix
(samples × SNPs), applies LW shrinkage via `cvCovEst::linearShrinkLWEst`,
normalises to correlation, zeroes negative eigenvalues, and counts.

```{r lw-faithful-helper}
## lw_shrink_lw: Python-faithful LW shrinkage for a genotype window
##
## @param win_geno  Numeric matrix (n_samples x p_snps) — raw genotypes,
##                  already mean-imputed (no NAs). Rows = samples, cols = SNPs.
## @return          Shrunk correlation matrix (p x p)
lw_shrink_lw <- function(win_geno) {
  p <- ncol(win_geno)
  if (p == 1L) return(matrix(1, 1, 1))

  ## Ledoit-Wolf 2004 shrunk covariance (from cvCovEst)
  shrunk_cov <- cvCovEst::linearShrinkLWEst(dat = win_geno)

  ## Handle degenerate case (all-identical columns)
  diag_vals <- diag(shrunk_cov)
  if (any(diag_vals <= 0)) {
    ## Fall back: all-ones correlation (perfect LD)
    return(matrix(1, p, p))
  }

  ## Convert shrunk covariance to correlation
  d_inv     <- diag(1 / sqrt(diag_vals))
  shrunk_cor <- d_inv %*% shrunk_cov %*% d_inv

  ## Enforce exact unit diagonal (floating point safety)
  diag(shrunk_cor) <- 1

  return(shrunk_cor)
}

## eigenMT_gene_lw: per-gene M_eff using the LW-faithful shrinkage
##
## Identical logic to eigenMT_gene() but uses lw_shrink_lw() instead of
## lw_shrink_cor(), extracting the raw genotype window from geno_mat_t
## (samples x SNPs matrix already in memory from Section 6).
##
## @param geno_mat_t  Numeric matrix (N_SAMPLES x N_SNPS) — full imputed
##                    genotype matrix, samples on rows.
## @param snp_indices Integer vector of column indices for cis-SNPs of this gene
## @param var_thresh  Variance threshold (default 0.99)
## @param window      Window size (default 200)
## @return Integer M_eff
eigenMT_gene_lw <- function(geno_mat_t, snp_indices,
                             var_thresh = 0.99, window = 200L) {
  M <- length(snp_indices)
  if (M == 0L) return(0L)
  if (M == 1L) return(1L)

  snp_indices <- sort(snp_indices)
  m_eff <- 0L
  start <- 1L

  while (start <= M) {
    stop     <- min(start + window - 1L, M)
    win_size <- stop - start + 1L

    if (win_size == 1L) {
      m_eff <- m_eff + 1L
      break
    }

    win_idx  <- snp_indices[start:stop]
    win_geno <- geno_mat_t[, win_idx, drop = FALSE]  # n x p, in memory

    shrunk_cor <- lw_shrink_lw(win_geno)

    eigs <- eigen(shrunk_cor, symmetric = TRUE, only.values = TRUE)$values
    eigs[eigs < 0] <- 0

    m_eff <- m_eff + count_eigenvalues(eigs, var_thresh)
    start <- stop + 1L
  }

  return(m_eff)
}
```

# 10. Run All Three R Implementations

```{r run-r-all}
CIS_DIST   <- 1e6
VAR_THRESH <- 0.99
EIGEN_WIN  <- 200L

r_results <- lapply(seq_len(nrow(features_coord)), function(i) {
  gene_row <- features_coord[i, ]

  cis_result <- get_cis_snps(
    bigsnp     = bigsnp_obj,
    pheno_chr   = gene_row$chromosome,
    pheno_start = gene_row$start,
    pheno_end   = gene_row$end,
    cis_window = CIS_DIST
  )
  idx    <- cis_result$indices
  n_cis  <- length(idx)

  ## Method A: current bigQTL (OAS on correlation)
  m_eff_oas <- eigenMT_gene(
    bigsnp      = bigsnp_obj,
    snp_indices = idx,
    ind.row     = ind.row,
    var_thresh  = VAR_THRESH,
    window      = EIGEN_WIN
  )

  ## Method B: Python-faithful (LW on raw genotype submatrix)
  m_eff_lw <- eigenMT_gene_lw(
    geno_mat_t  = geno_mat_t,
    snp_indices = idx,
    var_thresh  = VAR_THRESH,
    window      = EIGEN_WIN
  )

  data.frame(
    gene        = gene_row$feature_name,
    n_cis_snps  = n_cis,
    m_eff_oas   = m_eff_oas,
    m_eff_lw    = m_eff_lw,
    stringsAsFactors = FALSE
  )
})

r_df <- do.call(rbind, r_results)
message(sprintf("R eigenMT: %d genes", nrow(r_df)))
print(head(r_df))
```

# 11. Unit Tests

```{r unit-tests}
n_failures <- 0L

## 11a. count_eigenvalues vs Python find_num_eigs ----------------------------
py_find_num_eigs <- function(eigenvalues, variance, var_thresh) {
  eigenvalues <- sort(eigenvalues, decreasing = TRUE)
  running_sum <- 0; counter <- 0L
  while (running_sum < variance * var_thresh) {
    running_sum <- running_sum + eigenvalues[counter + 1L]
    counter     <- counter + 1L
  }
  counter
}
test_eigs <- c(3.7, 2.3, 2, 1.5, 1.1, 1, 0.8, 0.5, 0.4, 0.3, 0.2, 0.1, 0.05, 0.05)
r_c  <- count_eigenvalues(test_eigs, 0.99)
py_c <- py_find_num_eigs(test_eigs, length(test_eigs), 0.99)
if (r_c != py_c) {
  n_failures <- n_failures + 1L
  message(sprintf("  FAIL count_eigenvalues: R=%d Python=%d", r_c, py_c))
} else {
  message(sprintf("  PASS count_eigenvalues (both = %d)", r_c))
}

## 11b. big_cor vs base cor() ------------------------------------------------
test_cols <- seq_len(min(30L, N_SNPS))
base_cor  <- cor(geno_mat_t[, test_cols])
fbm_cor   <- bigstatsr::big_cor(geno_fbm, ind.row = ind.row, ind.col = test_cols)[]
frob      <- sqrt(sum((base_cor - fbm_cor)^2))
if (frob >= 1e-6) {
  n_failures <- n_failures + 1L
  message(sprintf("  FAIL big_cor vs cor(): Frob=%.2e", frob))
} else {
  message(sprintf("  PASS big_cor vs cor() (Frob=%.2e)", frob))
}

## 11c. lw_shrink_lw sanity ---------------------------------------------------
test_win <- geno_mat_t[, seq_len(min(20L, N_SNPS))]
lw_cor   <- lw_shrink_lw(test_win)
eigs_lw  <- eigen(lw_cor, symmetric = TRUE, only.values = TRUE)$values
if (!all(eigs_lw >= -1e-10)) {
  n_failures <- n_failures + 1L; message("  FAIL lw_shrink_lw: not PSD")
} else if (!all(abs(diag(lw_cor) - 1) <= 1e-10)) {
  n_failures <- n_failures + 1L; message("  FAIL lw_shrink_lw: diagonal != 1")
} else {
  message("  PASS lw_shrink_lw (PSD, diagonal = 1)")
}

## 11d. OAS lw_shrink_cor sanity ---------------------------------------------
R_test <- cor(geno_mat_t[, seq_len(min(20L, N_SNPS))])
oas_cor <- lw_shrink_cor(R_test, N_SAMPLES)
eigs_oas <- eigen(oas_cor, symmetric = TRUE, only.values = TRUE)$values
if (!all(eigs_oas >= -1e-10)) {
  n_failures <- n_failures + 1L; message("  FAIL lw_shrink_cor (OAS): not PSD")
} else if (!all(abs(diag(oas_cor) - 1) <= 1e-10)) {
  n_failures <- n_failures + 1L; message("  FAIL lw_shrink_cor (OAS): diagonal != 1")
} else {
  message("  PASS lw_shrink_cor (OAS, PSD, diagonal = 1)")
}

## 11e. eigenMT_gene_lw edge cases -------------------------------------------
m1 <- eigenMT_gene_lw(geno_mat_t, snp_indices = 1L)
m0 <- eigenMT_gene_lw(geno_mat_t, snp_indices = integer(0))
if (m1 != 1L) { n_failures <- n_failures + 1L; message("  FAIL single-SNP LW") }
if (m0 != 0L) { n_failures <- n_failures + 1L; message("  FAIL zero-SNP LW")   }
if (m1 == 1L && m0 == 0L) message("  PASS eigenMT_gene_lw edge cases")

message(sprintf("\nUnit tests: %d failure(s)", n_failures))
```

# 12. Full Comparison: Python vs OAS vs LW

```{r compare}
## Identify gene column in Python output
py_gene_col <- intersect(colnames(python_results),
                         c("gene", "geneid", "phenotype_id"))
if (length(py_gene_col) == 0L)
  stop("Cannot identify gene col in Python output: ",
       paste(colnames(python_results), collapse = ", "))
py_gene_col <- py_gene_col[1L]

comparison <- do.call(rbind, lapply(r_df$gene, function(gene) {
  r_row  <- r_df[r_df$gene == gene, ]
  py_row <- python_results[python_results[[py_gene_col]] == gene, ]
  if (nrow(r_row) == 0L || nrow(py_row) == 0L) return(NULL)
  m_py  <- py_row$TESTS[1L]
  m_oas <- r_row$m_eff_oas[1L]
  m_lw  <- r_row$m_eff_lw[1L]
  data.frame(
    gene       = gene,
    n_cis_snps = r_row$n_cis_snps[1L],
    m_eff_py   = m_py,
    m_eff_oas  = m_oas,
    m_eff_lw   = m_lw,
    rel_diff_oas = abs(m_oas - m_py) / max(m_py, 1L),
    rel_diff_lw  = abs(m_lw  - m_py) / max(m_py, 1L),
    stringsAsFactors = FALSE
  )
}))

message(sprintf("Comparison: %d genes", nrow(comparison)))
print(comparison)

message(sprintf(
  "\nM_eff means — Python: %.1f | OAS: %.1f | LW: %.1f",
  mean(comparison$m_eff_py),
  mean(comparison$m_eff_oas),
  mean(comparison$m_eff_lw)
))
message(sprintf(
  "Relative diff vs Python — OAS: median=%.0f%% max=%.0f%% | LW: median=%.0f%% max=%.0f%%",
  100 * median(comparison$rel_diff_oas), 100 * max(comparison$rel_diff_oas),
  100 * median(comparison$rel_diff_lw),  100 * max(comparison$rel_diff_lw)
))

MEFF_TOL <- 0.20
bad_lw <- comparison[comparison$rel_diff_lw > MEFF_TOL, ]
if (nrow(bad_lw) > 0L) {
  message(sprintf("WARNING: %d gene(s) with LW M_eff diff > %d%%:",
                  nrow(bad_lw), as.integer(MEFF_TOL * 100)))
  print(bad_lw[, c("gene", "n_cis_snps", "m_eff_py", "m_eff_lw", "rel_diff_lw")])
} else {
  message(sprintf("All %d genes: LW within %d%% of Python",
                  nrow(comparison), as.integer(MEFF_TOL * 100)))
}
```

# 13. Clean Up

```{r cleanup}
unlink(tmp_dir, recursive = TRUE)
message("Temp files removed.")
```
