---
title: "eigenMT Validation: R vs Python"
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, cache = TRUE)
```

# Overview

This vignette validates the R implementation of eigenMT (implemented in
`bigQTL`) against the **original** Python eigenMT
([joed3/eigenMT](https://github.com/joed3/eigenMT)) by running both on the
real example dataset shipped with that repository and diagnosing the sources
of any difference.

## Known implementation differences (updated after audit)

| # | Area | Python | R `bigQTL` (current, `"basic"`) | Severity |
|---|------|--------|---------------------------------|----------|
| 1 | **Shrinkage estimator — input** | `LedoitWolf.fit()` on raw centred genotype matrix (n × p); produces shrunk **covariance**, then normalised to correlation | `lw_shrink_geno()`: same approach — LW 2004 on biased sample covariance, then `cov2cor` | 🟢 Equivalent |
| 2 | **Shrinkage formula** | Ledoit-Wolf 2004 analytic formula | Same analytical formula | 🟢 Equivalent |
| 3 | **Missing genotype imputation** | `NA` → column mean before any computation | Mean-imputed before building FBM in this vignette | 🟢 Equivalent (when imputed) |
| 4 | **SNP ordering** | By genomic position | By FBM column index | 🟡 Minor (equivalent when FBM is position-sorted) |
| 5 | `count_eigenvalues` normalisation | Window size `m` as `variance` | `sum(eigenvalues)`; equivalent since `trace(cor) == p` | 🟢 Equivalent |


## Shrinkage methods in `bigQTL`

`bigQTL::eigenMT_gene()` and `bigQTL::eigenMT_batch()` support two shrinkage
methods via the `shrinkage_method` argument:

| Method | Shrinkage estimator | Package dependency | Expected match to Python |
|--------|--------------------|--------------------|--------------------------|
| `"basic"` (default) | `lw_shrink_geno()`: LW 2004 analytical on biased genotype covariance | None (base R only) | Very close |
| `"nlshrink"` | `nlshrink::linshrink_cov()` — ports [ImmuneAxisa/eigenMT](https://github.com/ImmuneAxisa/eigenMT) | `nlshrink` (Suggests) | Equivalent to Python |


## Prerequisites

Note: it is crucial to enforce  `scikit-learn < 1.2` to match the original eigenMT environment, otherwise the shrinkage silently fails for all windows and returning to one effective test per window. See [ImmuneAxisa/eigenMT](https://github.com/ImmuneAxisa/eigenMT) for more details.

```{r conda-setup}
library(reticulate)

if (!any("eigenMT" %in% conda_list()[["name"]])) {
  conda_create(
    "eigenMT",
    packages = c("numpy", "scipy", "scikit-learn<1.2", "pandas"),
    python_version = "3.8"
  )
}
```

```{r load-pkg}
library(bigQTL)
library(bigstatsr)
library(bigsnpr)
# if you want to use the optional "nlshrink" method, install and load nlshrink:
if(!requireNamespace("nlshrink", quietly = TRUE)) {
  message("Optional 'nlshrink' package not found. Installing")
  install.packages("nlshrink")
}
library(nlshrink)
```

# Data retrieval and processing steps

## Download and Decompress the Example Data

The example data contains a minimal example to eQTL results dataset run with the MatrixeQTL R package.

```{r download-example}
tmp_dir <- file.path(tempdir(), "eigenMT_validation")
dir.create(tmp_dir, recursive = TRUE, showWarnings = FALSE)

tgz_url  <- paste0(
  "https://github.com/joed3/eigenMT/raw/",
  "f91e2837d34403cff48c9a248a7d4fb3fac98541/example.tar.gz"
)
tgz_file <- file.path(tmp_dir, "example.tar.gz")
utils::download.file(tgz_url, destfile = tgz_file, mode = "wb", quiet = FALSE)
message(sprintf("Downloaded %.2f MB", file.info(tgz_file)$size / 1e6))

utils::untar(tgz_file, exdir = tmp_dir)
example_dir <- file.path(tmp_dir, "example")
stopifnot(dir.exists(example_dir))
message("Example files: ", paste(list.files(example_dir), collapse = ", "))
```

## Download the Original Python Script

```{r download-python}
py_script_url <- paste0(
  "https://raw.githubusercontent.com/joed3/eigenMT/",
  "f91e2837d34403cff48c9a248a7d4fb3fac98541/eigenMT.py"
)
py_script_file <- file.path(tmp_dir, "eigenMT.py")
utils::download.file(py_script_url, destfile = py_script_file, quiet = FALSE)
message("Downloaded eigenMT.py")
```

## Read the Example Data

```{r read-data}
geno_raw <- read.table(file.path(example_dir, "genotypes.txt"),
                       header = TRUE, sep = "\t",
                       check.names = FALSE, stringsAsFactors = FALSE)
message(sprintf("Genotype matrix: %d SNPs x %d samples",
                nrow(geno_raw), ncol(geno_raw) - 1L))

gen_pos <- read.table(file.path(example_dir, "gen.positions.txt"),
                      header = TRUE, sep = "\t", stringsAsFactors = FALSE)
phe_pos <- read.table(file.path(example_dir, "phe.positions.txt"),
                      header = TRUE, sep = "\t", stringsAsFactors = FALSE)
eqtl    <- read.table(file.path(example_dir, "cis.eqtls.txt"),
                      header = TRUE, sep = "\t", stringsAsFactors = FALSE)

message(sprintf("SNPs: %d | Genes: %d | eQTL rows: %d",
                nrow(gen_pos), nrow(phe_pos), nrow(eqtl)))
message(sprintf("eQTL file columns: %s", paste(colnames(eqtl), collapse = ", ")))
```

## Detect Chromosome

```{r detect-chrom}
CHROM <- as.character(unique(gen_pos[[2L]])[1L])
message(sprintf("Using chromosome: %s", CHROM))
```

## Run Python eigenMT

```{r run-python}
py_out_file <- file.path(tmp_dir, "eigenMT_output_py.txt")

reticulate::conda_run2(
  "python",
  args = c(
    py_script_file,
    "--CHROM",      CHROM,
    "--QTL",        file.path(example_dir, "cis.eqtls.txt"),
    "--GEN",        file.path(example_dir, "genotypes.txt"),
    "--GENPOS",     file.path(example_dir, "gen.positions.txt"),
    "--PHEPOS",     file.path(example_dir, "phe.positions.txt"),
    "--OUT",        py_out_file,
    "--cis_dist",   "1000000",
    "--var_thresh", "0.99",
    "--window",     "200"
  ),
  envname = "eigenMT"
)

python_results <- read.table(py_out_file, header = TRUE,
                             sep = "\t", stringsAsFactors = FALSE)
message(sprintf("Python eigenMT: %d genes", nrow(python_results)))
print(head(python_results))
```

## Build bigSNP Object

To match the data structure expected by `bigQTL`, we build a `bigSNP` object from the raw genotype matrix, ensuring that the SNPs are ordered by genomic position (to mirror Python's approach) and that missing values are imputed with the column mean before building the FBM (to mirror Python's imputation).

```{r build-bigsnp}
sample_ids <- colnames(geno_raw)[-1L]
N_SAMPLES  <- length(sample_ids)
snp_ids    <- as.character(geno_raw[[1L]])
geno_mat   <- as.matrix(geno_raw[, -1L, drop = FALSE])
storage.mode(geno_mat) <- "double"

## Samples x SNPs
geno_mat_t <- t(geno_mat)
N_SNPS     <- ncol(geno_mat_t)

## SNP position lookup
snp_order <- match(snp_ids, as.character(gen_pos[[1L]]))
if (any(is.na(snp_order)))
  warning(sum(is.na(snp_order)), " SNP IDs not found in gen.positions.txt")
chr_vec <- as.character(gen_pos[[2L]][snp_order])
pos_vec <- as.numeric(gen_pos[[3L]][snp_order])

## Impute NA -> column mean (mirrors Python)
for (j in seq_len(N_SNPS)) {
  col_j <- geno_mat_t[, j]
  na_m  <- is.na(col_j)
  if (any(na_m))
    geno_mat_t[na_m, j] <- mean(col_j, na.rm = TRUE)
}

## FBM
fbm_file <- file.path(tmp_dir, "geno")
geno_fbm <- bigstatsr::FBM(
  nrow = N_SAMPLES, ncol = N_SNPS, type = "double",
  backingfile = fbm_file, create_bk = TRUE
)
geno_fbm[] <- geno_mat_t

snp_map <- data.frame(
  chromosome = chr_vec, marker.ID = snp_ids,
  genetic.dist = 0, physical.pos = pos_vec,
  allele1 = "A", allele2 = "C",
  stringsAsFactors = FALSE
)
snp_fam <- data.frame(
  family.ID = sample_ids, sample.ID = sample_ids,
  paternal.ID = 0L, maternal.ID = 0L,
  sex = 0L, affection = 0L,
  stringsAsFactors = FALSE
)
bigsnp_obj <- structure(
  list(genotypes = geno_fbm, map = snp_map, fam = snp_fam),
  class = "bigSNP"
)
ind.row <- seq_len(N_SAMPLES)
message(sprintf("bigSNP: %d samples x %d SNPs", N_SAMPLES, N_SNPS))
```


# Implementation comparions

## Gene Coordinates

```{r build-features}
genes_in_eqtl  <- unique(eqtl[[2L]])
features_coord <- data.frame(
  feature_name = as.character(phe_pos[[1L]]),
  chromosome   = as.character(phe_pos[[2L]]),
  start        = as.numeric(phe_pos[[3L]]),
  end          = as.numeric(phe_pos[[4L]]),
  stringsAsFactors = FALSE
)
features_coord <- features_coord[
  features_coord$chromosome == CHROM &
    features_coord$feature_name %in% genes_in_eqtl, ]

## Sub-sample for faster article testing (first 20 genes)
N_GENES_SAMPLE <- 20L
if (nrow(features_coord) > N_GENES_SAMPLE) {
  features_coord <- features_coord[seq_len(N_GENES_SAMPLE), ]
  message(sprintf("Sub-sampled to first %d genes for article testing", N_GENES_SAMPLE))
}
message(sprintf("Genes to process: %d", nrow(features_coord)))
```

## Shrinkage Diagnostic — Side-by-Side on One Window

Inspect the shrinkage intensity of the current `bigQTL` `"basic"` approach
vs the Python-faithful cvCovEst approach on a small test window.

```{r shrinkage-diagnostic}
## Take the first 50 SNPs as a test window
diag_cols <- seq_len(min(50L, N_SNPS))
win_geno  <- geno_mat_t[, diag_cols]   # n x p (samples x SNPs)

## --- Method A: new bigQTL "basic" (lw_shrink_geno) ---
shrunk_basic <- lw_shrink_geno(win_geno)
eigs_basic   <- eigen(shrunk_basic, symmetric = TRUE, only.values = TRUE)$values
eigs_basic[eigs_basic < 0] <- 0
m_eff_basic  <- count_eigenvalues(eigs_basic, 0.99)

## Historical reference: show what OAS-on-correlation gives
R_raw <- cor(win_geno)
p     <- ncol(win_geno)
n     <- nrow(win_geno)
trR2  <- sum(R_raw^2)
num   <- (1 - 2/p) * trR2 + p^2
den   <- (n + 1 - 2/p) * (trR2 - p)
alpha_oas <- max(0, min(1, num / den))
message(sprintf(
  "Historical OAS-on-correlation alpha: %.6f  (near 0 = almost no shrinkage)",
  alpha_oas
))
message(sprintf(
  "New lw_shrink_geno M_eff (p=%d window): %d  (cf. OAS which would give ~%d)",
  p, m_eff_basic, count_eigenvalues(pmax(eigen(R_raw, symmetric = TRUE,
    only.values = TRUE)$values, 0), 0.99)
))

## --- Method B: Python-faithful via cvCovEst::linearShrinkLWEst ---
if (requireNamespace("cvCovEst", quietly = TRUE)) {
  lw_cov  <- cvCovEst::linearShrinkLWEst(dat = win_geno)
  d_inv   <- diag(1 / sqrt(diag(lw_cov)))
  lw_cor  <- d_inv %*% lw_cov %*% d_inv
  diag(lw_cor) <- 1
  eigs_lw <- eigen(lw_cor, symmetric = TRUE, only.values = TRUE)$values
  eigs_lw[eigs_lw < 0] <- 0
  m_eff_lw_cv <- count_eigenvalues(eigs_lw, 0.99)
  message(sprintf("cvCovEst LW M_eff (p=%d window): %d", p, m_eff_lw_cv))
} else {
  message("Install cvCovEst for the Python-faithful diagnostic: install.packages('cvCovEst')")
}
```


## Run All R Implementations

2 methods are compared against Python:

- **Method A** (`"basic"`): new `bigQTL` default — `lw_shrink_geno()` (LW 2004 on biased
  genotype covariance, no extra dependencies)
- **Method B** (`"nlshrink"`): new `bigQTL` optional — `nlshrink::linshrink_cov()`
  (requires `nlshrink`)

```{r run-r-all}
CIS_DIST   <- 1e6
VAR_THRESH <- 0.99
EIGEN_WIN  <- 200L


r_results <- lapply(seq_len(nrow(features_coord)), function(i) {
  gene_row <- features_coord[i, ]

  cis_result <- get_cis_snps(
    bigsnp      = bigsnp_obj,
    pheno_chr   = gene_row$chromosome,
    pheno_start = gene_row$start,
    pheno_end   = gene_row$end,
    cis_window  = CIS_DIST
  )
  idx   <- cis_result$indices
  n_cis <- length(idx)

  ## Method A: bigQTL basic (lw_shrink_geno, no extra deps)
  m_eff_basic <- eigenMT_gene(
    bigsnp           = bigsnp_obj,
    snp_indices      = idx,
    ind.row          = ind.row,
    var_thresh       = VAR_THRESH,
    window           = EIGEN_WIN,
    shrinkage_method = "basic"
  )

  ## Method B: bigQTL nlshrink (optional)

  m_eff_nlshrink <- eigenMT_gene(
    bigsnp           = bigsnp_obj,
    snp_indices      = idx,
    ind.row          = ind.row,
    var_thresh       = VAR_THRESH,
    window           = EIGEN_WIN,
    shrinkage_method = "nlshrink"
  )

  data.frame(
    gene           = gene_row$feature_name,
    n_cis_snps     = n_cis,
    m_eff_basic    = m_eff_basic,
    m_eff_nlshrink = m_eff_nlshrink,
    stringsAsFactors = FALSE
  )
})

r_df <- do.call(rbind, r_results)
message(sprintf("R eigenMT: %d genes", nrow(r_df)))
print(head(r_df))
```


## Full Comparison: Python vs R Methods

```{r compare}
## Identify gene column in Python output
py_gene_col <- intersect(colnames(python_results),
                         c("gene", "geneid", "phenotype_id"))
if (length(py_gene_col) == 0L)
  stop("Cannot identify gene col in Python output: ",
       paste(colnames(python_results), collapse = ", "))
py_gene_col <- py_gene_col[1L]

comparison <- do.call(rbind, lapply(r_df$gene, function(gene) {
  r_row  <- r_df[r_df$gene == gene, ]
  py_row <- python_results[python_results[[py_gene_col]] == gene, ]
  if (nrow(r_row) == 0L || nrow(py_row) == 0L) return(NULL)
  m_py  <- py_row$TESTS[1L]
  m_b   <- r_row$m_eff_basic[1L]
  m_nls <- r_row$m_eff_nlshrink[1L]
  data.frame(
    gene              = gene,
    n_cis_snps        = r_row$n_cis_snps[1L],
    m_eff_py          = m_py,
    m_eff_basic       = m_b,
    m_eff_nlshrink    = m_nls,
    rel_diff_basic    = abs(m_b   - m_py) / max(m_py, 1L),
    rel_diff_nlshrink = if (!is.na(m_nls)) abs(m_nls - m_py) / max(m_py, 1L) else NA_real_,
    stringsAsFactors = FALSE
  )
}))

message(sprintf("Comparison: %d genes", nrow(comparison)))
print(comparison)

message(sprintf(
  "\nM_eff means — Python: %.1f | basic: %.1f | nlshrink: %s | lw (cvCovEst): %s",
  mean(comparison$m_eff_py),
  mean(comparison$m_eff_basic),
  if (!all(is.na(comparison$m_eff_nlshrink)))
    sprintf("%.1f", mean(comparison$m_eff_nlshrink, na.rm = TRUE)) else "n/a",
  if (!all(is.na(comparison$m_eff_lw)))
    sprintf("%.1f", mean(comparison$m_eff_lw, na.rm = TRUE)) else "n/a"
))

message(sprintf(
  "Relative diff vs Python — basic: median=%.0f%% max=%.0f%%",
  100 * median(comparison$rel_diff_basic, na.rm = TRUE),
  100 * max(comparison$rel_diff_basic, na.rm = TRUE)
))
if (!all(is.na(comparison$rel_diff_nlshrink)))
  message(sprintf(
    "                       nlshrink: median=%.0f%% max=%.0f%%",
    100 * median(comparison$rel_diff_nlshrink, na.rm = TRUE),
    100 * max(comparison$rel_diff_nlshrink, na.rm = TRUE)
  ))
if (!all(is.na(comparison$rel_diff_lw)))
  message(sprintf(
    "                         lw (cvCovEst): median=%.0f%% max=%.0f%%",
    100 * median(comparison$rel_diff_lw, na.rm = TRUE),
    100 * max(comparison$rel_diff_lw, na.rm = TRUE)
  ))

MEFF_TOL <- 0.20
bad_basic <- comparison[comparison$rel_diff_basic > MEFF_TOL, ]
if (nrow(bad_basic) > 0L) {
  message(sprintf("WARNING: %d gene(s) with basic M_eff diff > %d%%:",
                  nrow(bad_basic), as.integer(MEFF_TOL * 100)))
  print(bad_basic[, c("gene", "n_cis_snps", "m_eff_py", "m_eff_basic",
                       "rel_diff_basic")])
} else {
  message(sprintf("All %d genes: basic within %d%% of Python",
                  nrow(comparison), as.integer(MEFF_TOL * 100)))
}
```

# 13. Clean Up

```{r cleanup}
unlink(tmp_dir, recursive = TRUE)
message("Temp files removed.")
```
