---
title: "eigenMT Validation: R vs Python"
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Overview

This vignette validates the R implementation of eigenMT (implemented in
`bigQTL`) against the **original** Python eigenMT
([joed3/eigenMT](https://github.com/joed3/eigenMT)) by running both on the
real example dataset shipped with that repository.

The example data (`example.tar.gz`) contains a real Matrix-eQTL run on
chromosome 1: genotype matrix, SNP positions, phenotype positions, and
cis-eQTL results. We download the archive at run-time, decompress it, and
feed the same files to both implementations.

All code chunks use `eval = FALSE` and are intended to be run interactively.

## Known implementation differences

After a careful comparison of the original Python `eigenMT.py` and
`R/eigenMT.R` in this package, the following discrepancies were identified:

| # | Area | Python | R (`bigQTL`) | Severity |
|---|------|--------|--------------|----------|
| 1 | **Shrinkage estimator** | `sklearn.covariance.LedoitWolf` fitted on the raw genotype matrix (SNPs × samples); shrinkage intensity estimated analytically per Ledoit & Wolf 2004 | OAS closed-form formula (Chen et al. 2010) applied to a pre-computed sample *correlation* matrix | 🔴 Major — different alpha values expected |
| 2 | **Missing genotype imputation** | `NA` → `-1` → column-mean imputation before computing correlation | No explicit imputation; `bigstatsr::big_cor()` will propagate `NA` | 🟠 Moderate |
| 3 | **SNP ordering** | SNPs sorted by genomic *position* (keys of `gen_dict`) | SNPs sorted by FBM *column index*; equivalent only if FBM is already position-sorted | 🟡 Minor |
| 4 | `count_eigenvalues` normalisation | Uses window size `m` as `variance` argument | Uses `sum(eigenvalues)`; equivalent because `trace(shrunk_cor) == p` | 🟢 Equivalent |
| 5 | Window traversal | Disjoint windows, `break` on single-SNP remainder | Disjoint windows, same `break` logic | 🟢 Equivalent |

The most important discrepancy is **#1**: sklearn's `LedoitWolf` and the OAS
formula are both shrinkage estimators targeting the identity, but they compute
the shrinkage intensity differently, leading to different effective test counts.

## Prerequisites

You need a conda environment named `"eigenMT"` with the required Python
packages. Create it once with:

```{r conda-setup}
library(reticulate)
conda_create(
  "eigenMT",
  packages = c("numpy", "scipy", "scikit-learn", "pandas"),
  python_version = "3.8"
)
```

Then load the packages:

```{r load-pkg}
library(bigQTL)
library(bigstatsr)
library(bigsnpr)
library(reticulate)
```

# 1. Download and Decompress the Example Data

The example archive is fetched directly from the joed3/eigenMT GitHub
repository at a pinned commit and decompressed into a temporary directory.

```{r download-example}
tmp_dir <- file.path(tempdir(), "eigenMT_validation")
dir.create(tmp_dir, recursive = TRUE, showWarnings = FALSE)

## Download the archive
tgz_url  <- paste0(
  "https://github.com/joed3/eigenMT/raw/",
  "f91e2837d34403cff48c9a248a7d4fb3fac98541/example.tar.gz"
)
tgz_file <- file.path(tmp_dir, "example.tar.gz")

utils::download.file(tgz_url, destfile = tgz_file, mode = "wb",
                     quiet = FALSE)
message(sprintf("Downloaded %.1f MB to %s",
                file.info(tgz_file)$size / 1e6, tgz_file))

## Decompress
utils::untar(tgz_file, exdir = tmp_dir)

## The archive unpacks to an "example/" sub-directory
example_dir <- file.path(tmp_dir, "example")
stopifnot(dir.exists(example_dir))
message("Example data files:")
print(list.files(example_dir))
```

After decompression the `example/` folder contains:

- `genotypes.txt`     — SNPs × samples genotype matrix
- `gen.positions.txt` — SNP id / chr / position
- `phe.positions.txt` — gene id / chr / start / end
- `cis.eqtls.txt`     — Matrix-eQTL output (one row per cis SNP–gene pair)

# 2. Download the Original Python Script

We fetch `eigenMT.py` directly from the same pinned commit, so there is no
embedded or simplified copy — we run the real thing.

```{r download-python}
py_script_url <- paste0(
  "https://raw.githubusercontent.com/joed3/eigenMT/",
  "f91e2837d34403cff48c9a248a7d4fb3fac98541/eigenMT.py"
)
py_script_file <- file.path(tmp_dir, "eigenMT.py")

utils::download.file(py_script_url, destfile = py_script_file,
                     quiet = FALSE)
message("Downloaded eigenMT.py")
```

# 3. Inspect the Example Data

```{r inspect-data}
## Genotype matrix: first column is the SNP id
geno_raw <- read.table(file.path(example_dir, "genotypes.txt"),
                       header = TRUE, sep = "\t",
                       check.names = FALSE, stringsAsFactors = FALSE)

message(sprintf("Genotype matrix: %d SNPs x %d samples (+1 id col)",
                nrow(geno_raw), ncol(geno_raw) - 1L))

## SNP positions: columns  snp / chr / pos
gen_pos <- read.table(file.path(example_dir, "gen.positions.txt"),
                      header = TRUE, sep = "\t",
                      stringsAsFactors = FALSE)

message(sprintf("SNP positions: %d rows; chromosomes: %s",
                nrow(gen_pos),
                paste(unique(gen_pos[[2L]]), collapse = ", ")))

## Phenotype positions: columns  geneid / chr / s1 / s2
phe_pos <- read.table(file.path(example_dir, "phe.positions.txt"),
                      header = TRUE, sep = "\t",
                      stringsAsFactors = FALSE)
message(sprintf("Gene positions: %d genes", nrow(phe_pos)))

## cis-eQTL results (Matrix-eQTL format)
eqtl <- read.table(file.path(example_dir, "cis.eqtls.txt"),
                   header = TRUE, sep = "\t",
                   stringsAsFactors = FALSE)
message(sprintf("cis-eQTL tests: %d rows; columns: %s",
                nrow(eqtl), paste(colnames(eqtl), collapse = ", ")))
```

# 4. Run Python eigenMT

We run the original `eigenMT.py` via `reticulate::conda_run2()` using the
`"eigenMT"` conda environment. `--CHROM 1` matches the example data.

```{r run-python}
py_out_file <- file.path(tmp_dir, "eigenMT_output_py.txt")

reticulate::conda_run2(
  "python",
  args = c(
    py_script_file,
    "--CHROM",      "19",
    "--QTL",        file.path(example_dir, "cis.eqtls.txt"),
    "--GEN",        file.path(example_dir, "genotypes.txt"),
    "--GENPOS",     file.path(example_dir, "gen.positions.txt"),
    "--PHEPOS",     file.path(example_dir, "phe.positions.txt"),
    "--OUT",        py_out_file,
    "--cis_dist",   "1000000",
    "--var_thresh", "0.99",
    "--window",     "200"
  ),
  envname = "eigenMT"
)

python_results <- read.table(py_out_file, header = TRUE,
                             sep = "\t", stringsAsFactors = FALSE)
message(sprintf("Python eigenMT: %d genes processed", nrow(python_results)))
print(head(python_results))
```

# 5. Build a bigSNP Object from the Example Genotype Matrix

The R implementation works on FBM-backed genotype matrices (bigSNP objects).
We construct one from the same flat-file genotypes used by Python, applying
the same missing-value imputation (column mean) that Python performs.

```{r build-bigsnp}
## Sample IDs: columns 2..end of the genotype matrix
sample_ids <- colnames(geno_raw)[-1L]
N_SAMPLES  <- length(sample_ids)

## SNP ids and integer genotype matrix (SNPs × samples)
snp_ids  <- as.character(geno_raw[[1L]])
geno_mat <- as.matrix(geno_raw[, -1L, drop = FALSE])
storage.mode(geno_mat) <- "integer"

## Transpose to samples × SNPs for the FBM
geno_mat_t <- t(geno_mat)
N_SNPS     <- ncol(geno_mat_t)

## Match SNP positions from gen.positions.txt
## Column order: snp_id / chr / pos  (use positional indexing for robustness)
snp_order <- match(snp_ids, as.character(gen_pos[[1L]]))
if (any(is.na(snp_order))) {
  warning(sprintf(
    "%d SNP IDs in genotypes.txt not found in gen.positions.txt",
    sum(is.na(snp_order))
  ))
}
chr_vec <- as.character(gen_pos[[2L]][snp_order])
pos_vec <- as.numeric( gen_pos[[3L]][snp_order])

## Impute NA genotypes with per-SNP column means
## (mirrors Python: genos[genos == -1] = np.mean(genos[genos != -1]))
for (j in seq_len(N_SNPS)) {
  col_j   <- geno_mat_t[, j]
  na_mask <- is.na(col_j)
  if (any(na_mask)) {
    geno_mat_t[na_mask, j] <- as.integer(round(mean(col_j, na.rm = TRUE)))
  }
}

## Create FBM
fbm_file <- file.path(tmp_dir, "geno")
geno_fbm <- bigstatsr::FBM(
  nrow = N_SAMPLES, ncol = N_SNPS, type = "integer",
  backingfile = fbm_file,
  create_bk   = TRUE
)
geno_fbm[] <- geno_mat_t

## Assemble bigSNP-like object
snp_map <- data.frame(
  chromosome   = chr_vec,
  marker.ID    = snp_ids,
  genetic.dist = rep(0, N_SNPS),
  physical.pos = pos_vec,
  allele1      = rep("A", N_SNPS),
  allele2      = rep("C", N_SNPS),
  stringsAsFactors = FALSE
)

snp_fam <- data.frame(
  family.ID   = sample_ids,
  sample.ID   = sample_ids,
  paternal.ID = rep(0L, N_SAMPLES),
  maternal.ID = rep(0L, N_SAMPLES),
  sex         = rep(0L, N_SAMPLES),
  affection   = rep(0L, N_SAMPLES),
  stringsAsFactors = FALSE
)

bigsnp_obj <- structure(
  list(genotypes = geno_fbm, map = snp_map, fam = snp_fam),
  class = "bigSNP"
)
ind.row <- seq_len(N_SAMPLES)

message(sprintf("bigSNP object: %d samples, %d SNPs", N_SAMPLES, N_SNPS))
```

# 6. Build Gene Coordinates from phe.positions.txt

We restrict to chromosome 19 genes that appear in the cis-eQTL file,
mirroring Python's `--CHROM 19` filter.

```{r build-features}
## Second column of the eQTL file is the gene id
genes_in_eqtl <- unique(eqtl[[2L]])

## phe.positions.txt: geneid / chr / s1 / s2
features_coord <- data.frame(
  feature_name = as.character(phe_pos[[1L]]),
  chromosome   = as.character(phe_pos[[2L]]),
  start        = as.numeric(phe_pos[[3L]]),
  end          = as.numeric(phe_pos[[4L]]),
  stringsAsFactors = FALSE
)

## Keep chr1 genes that appear in the eQTL file
features_coord <- features_coord[
  features_coord$chromosome == "19" &
    features_coord$feature_name %in% genes_in_eqtl, ]

message(sprintf("Genes for R eigenMT: %d", nrow(features_coord)))
print(head(features_coord))
```

# 7. Run R eigenMT

```{r run-r-eigenmt}
CIS_DIST   <- 1e6
VAR_THRESH <- 0.99
EIGEN_WIN  <- 200L

r_results <- lapply(seq_len(nrow(features_coord)), function(i) {
  gene_row <- features_coord[i, ]

  cis_result <- get_cis_snps(
    bigsnp     = bigsnp_obj,
    gene_chr   = gene_row$chromosome,
    gene_start = gene_row$start,
    gene_end   = gene_row$end,
    cis_window = CIS_DIST
  )

  m_eff <- eigenMT_gene(
    bigsnp      = bigsnp_obj,
    snp_indices = cis_result$indices,
    ind.row     = ind.row,
    var_thresh  = VAR_THRESH,
    window      = EIGEN_WIN
  )

  data.frame(
    gene       = gene_row$feature_name,
    n_cis_snps = length(cis_result$indices),
    m_eff_r    = m_eff,
    stringsAsFactors = FALSE
  )
})

r_df <- do.call(rbind, r_results)
message(sprintf("R eigenMT: %d genes processed", nrow(r_df)))
print(head(r_df))
```

# 8. Unit Tests

These tests verify internal consistency of the R helpers, independent of
the Python comparison.

```{r unit-tests}
n_failures <- 0L

## 8a. count_eigenvalues vs reference Python find_num_eigs -----------------
py_find_num_eigs <- function(eigenvalues, variance, var_thresh) {
  eigenvalues <- sort(eigenvalues, decreasing = TRUE)
  running_sum <- 0
  counter     <- 0L
  while (running_sum < variance * var_thresh) {
    running_sum <- running_sum + eigenvalues[counter + 1L]
    counter     <- counter + 1L
  }
  counter
}

test_eigs <- c(3.7, 2.3, 2, 1.5, 1.1, 1, 0.8, 0.5, 0.4, 0.3, 0.2, 0.1, 0.05, 0.05)
r_count  <- count_eigenvalues(test_eigs, 0.99)
py_count <- py_find_num_eigs(test_eigs, length(test_eigs), 0.99)
if (r_count != py_count) {
  n_failures <- n_failures + 1L
  message(sprintf("  FAIL count_eigenvalues: R=%d Python=%d", r_count, py_count))
} else {
  message(sprintf("  PASS count_eigenvalues (both = %d)", r_count))
}

## 8b. big_cor vs base cor() ------------------------------------------------
test_cols <- seq_len(min(30L, N_SNPS))
base_cor  <- cor(geno_mat_t[, test_cols])
fbm_cor   <- bigstatsr::big_cor(geno_fbm, ind.row = ind.row,
                                ind.col = test_cols)[]
frob_diff <- sqrt(sum((base_cor - fbm_cor)^2))
if (frob_diff >= 1e-8) {
  n_failures <- n_failures + 1L
  message(sprintf("  FAIL big_cor vs cor(): Frobenius diff = %.2e", frob_diff))
} else {
  message(sprintf("  PASS big_cor vs cor() (Frobenius diff = %.2e)", frob_diff))
}

## 8c. lw_shrink_cor sanity checks ------------------------------------------
R_test <- cor(geno_mat_t[, seq_len(min(20L, N_SNPS))])
shrunk  <- lw_shrink_cor(R_test, N_SAMPLES)
eigs    <- eigen(shrunk, symmetric = TRUE, only.values = TRUE)$values
if (!all(eigs >= -1e-10)) {
  n_failures <- n_failures + 1L
  message("  FAIL lw_shrink_cor: not PSD")
} else if (!all(abs(diag(shrunk) - 1) <= 1e-10)) {
  n_failures <- n_failures + 1L
  message("  FAIL lw_shrink_cor: diagonal != 1")
} else {
  message("  PASS lw_shrink_cor (PSD, diagonal = 1)")
}

## 8d. eigenMT_gene edge cases ----------------------------------------------
m1 <- eigenMT_gene(bigsnp_obj, snp_indices = 1L,         ind.row = ind.row)
m0 <- eigenMT_gene(bigsnp_obj, snp_indices = integer(0), ind.row = ind.row)
if (m1 != 1L) { n_failures <- n_failures + 1L; message("  FAIL single-SNP M_eff") }
if (m0 != 0L) { n_failures <- n_failures + 1L; message("  FAIL zero-SNP M_eff")   }
if (m1 == 1L && m0 == 0L) message("  PASS eigenMT_gene edge cases")

message(sprintf("\nUnit tests: %d failure(s)", n_failures))
```

# 9. Compare R vs Python M_eff

We join on gene ID and compute the absolute and relative difference in M_eff.
Because the R and Python shrinkage estimators differ (OAS vs sklearn
LedoitWolf — see the **Known implementation differences** table at the top),
some divergence is expected. The tolerance is set to 20% to flag large
discrepancies; tighten this as the implementation is refined.

```{r compare}
MEFF_TOL <- 0.20   # 20% relative difference tolerance

## Robustly detect the gene column name in Python output
py_gene_col <- intersect(colnames(python_results),
                         c("gene", "geneid", "phenotype_id"))
if (length(py_gene_col) == 0L) {
  stop(
    "Cannot identify gene column in Python eigenMT output. Columns: ",
    paste(colnames(python_results), collapse = ", ")
  )
}
py_gene_col <- py_gene_col[1L]

comparison <- do.call(rbind, lapply(r_df$gene, function(gene) {
  r_row  <- r_df[r_df$gene == gene, ]
  py_row <- python_results[python_results[[py_gene_col]] == gene, ]
  if (nrow(r_row) == 0L || nrow(py_row) == 0L) return(NULL)
  m_r  <- r_row$m_eff_r[1L]
  m_py <- py_row$TESTS[1L]
  data.frame(
    gene       = gene,
    n_cis_snps = r_row$n_cis_snps[1L],
    m_eff_r    = m_r,
    m_eff_py   = m_py,
    abs_diff   = abs(m_r - m_py),
    rel_diff   = abs(m_r - m_py) / max(m_py, 1L),
    stringsAsFactors = FALSE
  )
}))

message(sprintf("Comparison table (%d genes):", nrow(comparison)))
print(comparison)

## Summary statistics
message(sprintf(
  "M_eff summary — R: mean=%.1f sd=%.1f | Python: mean=%.1f sd=%.1f",
  mean(comparison$m_eff_r),  sd(comparison$m_eff_r),
  mean(comparison$m_eff_py), sd(comparison$m_eff_py)
))
message(sprintf(
  "Relative diff — median=%.1f%%  max=%.1f%%",
  100 * median(comparison$rel_diff),
  100 * max(comparison$rel_diff)
))

## Flag genes that exceed the tolerance
bad <- comparison[comparison$rel_diff > MEFF_TOL, ]
if (nrow(bad) > 0L) {
  message(sprintf(
    "\nWARNING: %d gene(s) with M_eff relative diff > %d%%:",
    nrow(bad), as.integer(MEFF_TOL * 100)
  ))
  print(bad)
} else {
  message(sprintf(
    "\nAll %d genes within %d%% M_eff tolerance",
    nrow(comparison), as.integer(MEFF_TOL * 100)
  ))
}
```

# 10. Clean Up

```{r cleanup}
unlink(tmp_dir, recursive = TRUE)
message("Temp files removed.")
```
