---
title: "eigenMT Validation: R vs Python"
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
```

# Overview

This vignette validates the R implementation of eigenMT (implemented in
`bigQTL`) against the **original** Python eigenMT
([joed3/eigenMT](https://github.com/joed3/eigenMT)) by running both on the
real example dataset shipped with that repository and diagnosing the sources
of any difference.

## Known implementation differences (updated after audit)

| # | Area | Python | R `bigQTL` (current, `"basic"`) | Severity |
|---|------|--------|---------------------------------|----------|
| 1 | **Shrinkage estimator — input** | `LedoitWolf.fit()` on raw centred genotype matrix (n × p); produces shrunk **covariance**, then normalised to correlation | `lw_shrink_geno()`: same approach — LW 2004 on biased sample covariance, then `cov2cor` | 🟢 Equivalent |
| 2 | **Shrinkage formula** | Ledoit-Wolf 2004 analytic formula | Same analytical formula | 🟢 Equivalent |
| 3 | **Missing genotype imputation** | `NA` → column mean before any computation | Mean-imputed before building FBM in this vignette | 🟢 Equivalent (when imputed) |
| 4 | **SNP ordering** | By genomic position | By FBM column index | 🟡 Minor (equivalent when FBM is position-sorted) |
| 5 | `count_eigenvalues` normalisation | Window size `m` as `variance` | `sum(eigenvalues)`; equivalent since `trace(cor) == p` | 🟢 Equivalent |
| 6 | Window traversal | Disjoint, `break` on 1-SNP tail | Same | 🟢 Equivalent |

**Historical note:** The previous `bigQTL` implementation used the OAS (Chen
et al. 2010) formula applied directly to a pre-computed *correlation* matrix.
Because the correlation matrix has a fixed diagonal of 1 by construction,
`tr(R)^2/p = p` is constant regardless of off-diagonal structure, causing OAS
to systematically underestimate shrinkage (alpha ≈ 0 for typical window sizes
and sample sizes). This is fixed in the current implementation.

## Shrinkage methods in `bigQTL`

`bigQTL::eigenMT_gene()` and `bigQTL::eigenMT_batch()` support two shrinkage
methods via the `shrinkage_method` argument:

| Method | Shrinkage estimator | Package dependency | Expected match to Python |
|--------|--------------------|--------------------|--------------------------|
| `"basic"` (default) | `lw_shrink_geno()`: LW 2004 analytical on biased genotype covariance | None (base R only) | Very close |
| `"nlshrink"` | `nlshrink::linshrink_cov()` (LW 2017 non-parametric) — ports [ImmuneAxisa/eigenMT](https://github.com/ImmuneAxisa/eigenMT) | `nlshrink` (Suggests) | Closest to Python |

## Prerequisites

```{r conda-setup}
library(reticulate)

if (!any("eigenMT" %in% conda_list()[["name"]])) {
  conda_create(
    "eigenMT",
    packages = c("numpy", "scipy", "scikit-learn<1.2", "pandas"),
    python_version = "3.8"
  )
}
```

```{r load-pkg}
library(bigQTL)
library(bigstatsr)
library(bigsnpr)
# install.packages("nlshrink")   # optional: for shrinkage_method = "nlshrink"
# install.packages("cvCovEst")   # optional: for the Python-faithful diagnostic below
```

# 1. Download and Decompress the Example Data

```{r download-example}
tmp_dir <- file.path(tempdir(), "eigenMT_validation")
dir.create(tmp_dir, recursive = TRUE, showWarnings = FALSE)

tgz_url  <- paste0(
  "https://github.com/joed3/eigenMT/raw/",
  "f91e2837d34403cff48c9a248a7d4fb3fac98541/example.tar.gz"
)
tgz_file <- file.path(tmp_dir, "example.tar.gz")
utils::download.file(tgz_url, destfile = tgz_file, mode = "wb", quiet = FALSE)
message(sprintf("Downloaded %.2f MB", file.info(tgz_file)$size / 1e6))

utils::untar(tgz_file, exdir = tmp_dir)
example_dir <- file.path(tmp_dir, "example")
stopifnot(dir.exists(example_dir))
message("Example files: ", paste(list.files(example_dir), collapse = ", "))
```

# 2. Download the Original Python Script

```{r download-python}
py_script_url <- paste0(
  "https://raw.githubusercontent.com/joed3/eigenMT/",
  "f91e2837d34403cff48c9a248a7d4fb3fac98541/eigenMT.py"
)
py_script_file <- file.path(tmp_dir, "eigenMT.py")
utils::download.file(py_script_url, destfile = py_script_file, quiet = FALSE)
message("Downloaded eigenMT.py")
```

# 3. Read the Example Data

```{r read-data}
geno_raw <- read.table(file.path(example_dir, "genotypes.txt"),
                       header = TRUE, sep = "\t",
                       check.names = FALSE, stringsAsFactors = FALSE)
message(sprintf("Genotype matrix: %d SNPs x %d samples",
                nrow(geno_raw), ncol(geno_raw) - 1L))

gen_pos <- read.table(file.path(example_dir, "gen.positions.txt"),
                      header = TRUE, sep = "\t", stringsAsFactors = FALSE)
phe_pos <- read.table(file.path(example_dir, "phe.positions.txt"),
                      header = TRUE, sep = "\t", stringsAsFactors = FALSE)
eqtl    <- read.table(file.path(example_dir, "cis.eqtls.txt"),
                      header = TRUE, sep = "\t", stringsAsFactors = FALSE)

message(sprintf("SNPs: %d | Genes: %d | eQTL rows: %d",
                nrow(gen_pos), nrow(phe_pos), nrow(eqtl)))
message(sprintf("eQTL file columns: %s", paste(colnames(eqtl), collapse = ", ")))
```

# 4. Detect Chromosome

```{r detect-chrom}
CHROM <- as.character(unique(gen_pos[[2L]])[1L])
message(sprintf("Using chromosome: %s", CHROM))
```

# 5. Run Python eigenMT

```{r run-python}
py_out_file <- file.path(tmp_dir, "eigenMT_output_py.txt")

reticulate::conda_run2(
  "python",
  args = c(
    py_script_file,
    "--CHROM",      CHROM,
    "--QTL",        file.path(example_dir, "cis.eqtls.txt"),
    "--GEN",        file.path(example_dir, "genotypes.txt"),
    "--GENPOS",     file.path(example_dir, "gen.positions.txt"),
    "--PHEPOS",     file.path(example_dir, "phe.positions.txt"),
    "--OUT",        py_out_file,
    "--cis_dist",   "1000000",
    "--var_thresh", "0.99",
    "--window",     "200"
  ),
  envname = "eigenMT"
)

python_results <- read.table(py_out_file, header = TRUE,
                             sep = "\t", stringsAsFactors = FALSE)
message(sprintf("Python eigenMT: %d genes", nrow(python_results)))
print(head(python_results))
```

# 6. Build bigSNP Object

```{r build-bigsnp}
sample_ids <- colnames(geno_raw)[-1L]
N_SAMPLES  <- length(sample_ids)
snp_ids    <- as.character(geno_raw[[1L]])
geno_mat   <- as.matrix(geno_raw[, -1L, drop = FALSE])
storage.mode(geno_mat) <- "double"

## Samples x SNPs
geno_mat_t <- t(geno_mat)
N_SNPS     <- ncol(geno_mat_t)

## SNP position lookup
snp_order <- match(snp_ids, as.character(gen_pos[[1L]]))
if (any(is.na(snp_order)))
  warning(sum(is.na(snp_order)), " SNP IDs not found in gen.positions.txt")
chr_vec <- as.character(gen_pos[[2L]][snp_order])
pos_vec <- as.numeric(gen_pos[[3L]][snp_order])

## Impute NA -> column mean (mirrors Python)
for (j in seq_len(N_SNPS)) {
  col_j <- geno_mat_t[, j]
  na_m  <- is.na(col_j)
  if (any(na_m))
    geno_mat_t[na_m, j] <- mean(col_j, na.rm = TRUE)
}

## FBM
fbm_file <- file.path(tmp_dir, "geno")
geno_fbm <- bigstatsr::FBM(
  nrow = N_SAMPLES, ncol = N_SNPS, type = "double",
  backingfile = fbm_file, create_bk = TRUE
)
geno_fbm[] <- geno_mat_t

snp_map <- data.frame(
  chromosome = chr_vec, marker.ID = snp_ids,
  genetic.dist = 0, physical.pos = pos_vec,
  allele1 = "A", allele2 = "C",
  stringsAsFactors = FALSE
)
snp_fam <- data.frame(
  family.ID = sample_ids, sample.ID = sample_ids,
  paternal.ID = 0L, maternal.ID = 0L,
  sex = 0L, affection = 0L,
  stringsAsFactors = FALSE
)
bigsnp_obj <- structure(
  list(genotypes = geno_fbm, map = snp_map, fam = snp_fam),
  class = "bigSNP"
)
ind.row <- seq_len(N_SAMPLES)
message(sprintf("bigSNP: %d samples x %d SNPs", N_SAMPLES, N_SNPS))
```

# 7. Gene Coordinates

```{r build-features}
genes_in_eqtl  <- unique(eqtl[[2L]])
features_coord <- data.frame(
  feature_name = as.character(phe_pos[[1L]]),
  chromosome   = as.character(phe_pos[[2L]]),
  start        = as.numeric(phe_pos[[3L]]),
  end          = as.numeric(phe_pos[[4L]]),
  stringsAsFactors = FALSE
)
features_coord <- features_coord[
  features_coord$chromosome == CHROM &
    features_coord$feature_name %in% genes_in_eqtl, ]
message(sprintf("Genes to process: %d", nrow(features_coord)))
```

# 8. Shrinkage Diagnostic — Side-by-Side Comparison on One Window

Before running the full analysis, inspect the shrinkage behaviour of the
three approaches on a small window of real data so we can understand why
OAS-on-correlation fails.

```{r shrinkage-diagnostic}
## Take the first 50 SNPs as a test window
diag_cols <- seq_len(min(50L, N_SNPS))
win_geno  <- geno_mat_t[, diag_cols]   # n x p (samples x SNPs)

## --- Method A: OAS on correlation (current bigQTL) ---
R_raw <- cor(win_geno)
p     <- ncol(win_geno)
n     <- nrow(win_geno)
trR2  <- sum(R_raw^2)
num   <- (1 - 2/p) * trR2 + p^2
den   <- (n + 1 - 2/p) * (trR2 - p)
alpha_oas <- max(0, min(1, num / den))
message(sprintf("OAS-on-correlation alpha: %.6f  (near 0 = almost no shrinkage)", alpha_oas))

## --- Method B: LW 2004 on raw data (Python-faithful) ---
lw_cov   <- cvCovEst::linearShrinkLWEst(dat = win_geno)
## Convert shrunk covariance to correlation
d_inv    <- diag(1 / sqrt(diag(lw_cov)))
lw_cor   <- d_inv %*% lw_cov %*% d_inv
## The LW shrinkage intensity (b_n2/d_n2) can be read from the off-diagonal shrinkage:
## alpha_lw = 1 - (off-diag of lw_cov) / (off-diag of sample_cov)
sample_cov  <- cov(win_geno)
alpha_lw_approx <- 1 - mean(lw_cov[row(lw_cov) != col(lw_cov)] /
                             sample_cov[row(sample_cov) != col(sample_cov)])
message(sprintf("LW2004-on-raw-data effective shrinkage: ~%.6f", alpha_lw_approx))

## --- Compare eigenvalue distributions ---
eigs_unshrunk <- eigen(R_raw,         symmetric=TRUE, only.values=TRUE)$values
eigs_oas      <- eigen(lw_shrink_cor(R_raw, n), symmetric=TRUE, only.values=TRUE)$values
eigs_lw       <- eigen(lw_cor,        symmetric=TRUE, only.values=TRUE)$values
eigs_unshrunk[eigs_unshrunk < 0] <- 0
eigs_oas[eigs_oas < 0]           <- 0
eigs_lw[eigs_lw < 0]             <- 0

m_eff_unshrunk <- count_eigenvalues(eigs_unshrunk, 0.99)
m_eff_oas      <- count_eigenvalues(eigs_oas,      0.99)
m_eff_lw       <- count_eigenvalues(eigs_lw,       0.99)

## Sub-sample for faster article testing (first 20 genes)
N_GENES_SAMPLE <- 20L
if (nrow(features_coord) > N_GENES_SAMPLE) {
  features_coord <- features_coord[seq_len(N_GENES_SAMPLE), ]
  message(sprintf("Sub-sampled to first %d genes for article testing", N_GENES_SAMPLE))
}
message(sprintf("Genes to process: %d", nrow(features_coord)))
```

# 8. Shrinkage Diagnostic — Side-by-Side on One Window

Inspect the shrinkage intensity of the current `bigQTL` `"basic"` approach
vs the Python-faithful cvCovEst approach on a small test window.

```{r shrinkage-diagnostic}
## Take the first 50 SNPs as a test window
diag_cols <- seq_len(min(50L, N_SNPS))
win_geno  <- geno_mat_t[, diag_cols]   # n x p (samples x SNPs)

## --- Method A: new bigQTL "basic" (lw_shrink_geno) ---
shrunk_basic <- lw_shrink_geno(win_geno)
eigs_basic   <- eigen(shrunk_basic, symmetric = TRUE, only.values = TRUE)$values
eigs_basic[eigs_basic < 0] <- 0
m_eff_basic  <- count_eigenvalues(eigs_basic, 0.99)

## Historical reference: show what OAS-on-correlation gives
R_raw <- cor(win_geno)
p     <- ncol(win_geno)
n     <- nrow(win_geno)
trR2  <- sum(R_raw^2)
num   <- (1 - 2/p) * trR2 + p^2
den   <- (n + 1 - 2/p) * (trR2 - p)
alpha_oas <- max(0, min(1, num / den))
message(sprintf(
  "Historical OAS-on-correlation alpha: %.6f  (near 0 = almost no shrinkage)",
  alpha_oas
))
message(sprintf(
  "New lw_shrink_geno M_eff (p=%d window): %d  (cf. OAS which would give ~%d)",
  p, m_eff_basic, count_eigenvalues(pmax(eigen(R_raw, symmetric = TRUE,
    only.values = TRUE)$values, 0), 0.99)
))

## --- Method B: Python-faithful via cvCovEst::linearShrinkLWEst ---
if (requireNamespace("cvCovEst", quietly = TRUE)) {
  lw_cov  <- cvCovEst::linearShrinkLWEst(dat = win_geno)
  d_inv   <- diag(1 / sqrt(diag(lw_cov)))
  lw_cor  <- d_inv %*% lw_cov %*% d_inv
  diag(lw_cor) <- 1
  eigs_lw <- eigen(lw_cor, symmetric = TRUE, only.values = TRUE)$values
  eigs_lw[eigs_lw < 0] <- 0
  m_eff_lw_cv <- count_eigenvalues(eigs_lw, 0.99)
  message(sprintf("cvCovEst LW M_eff (p=%d window): %d", p, m_eff_lw_cv))
} else {
  message("Install cvCovEst for the Python-faithful diagnostic: install.packages('cvCovEst')")
}
```

# 9. Python-Faithful Local Helper (cvCovEst)

Defines `lw_shrink_lw()` and `eigenMT_gene_lw()` as validation-only helpers
that use `cvCovEst::linearShrinkLWEst` — the closest in-memory equivalent to
Python's `sklearn.covariance.LedoitWolf`.

```{r lw-faithful-helper}
## lw_shrink_lw: Python-faithful LW shrinkage for a genotype window
## Requires cvCovEst package.
lw_shrink_lw <- function(win_geno) {
  p <- ncol(win_geno)
  if (p == 1L) return(matrix(1, 1, 1))

  shrunk_cov <- cvCovEst::linearShrinkLWEst(dat = win_geno)

  diag_vals <- diag(shrunk_cov)
  if (any(diag_vals <= 0)) return(matrix(1, p, p))

  d_inv      <- diag(1 / sqrt(diag_vals))
  shrunk_cor <- d_inv %*% shrunk_cov %*% d_inv
  diag(shrunk_cor) <- 1
  return(shrunk_cor)
}

## eigenMT_gene_lw: per-gene M_eff using lw_shrink_lw (in-memory genotype matrix)
eigenMT_gene_lw <- function(geno_mat_t, snp_indices,
                             var_thresh = 0.99, window = 200L) {
  M <- length(snp_indices)
  if (M == 0L) return(0L)
  if (M == 1L) return(1L)

  snp_indices <- sort(snp_indices)
  m_eff <- 0L
  start <- 1L

  while (start <= M) {
    stop     <- min(start + window - 1L, M)
    win_size <- stop - start + 1L

    if (win_size == 1L) {
      m_eff <- m_eff + 1L
      break
    }

    win_idx  <- snp_indices[start:stop]
    win_geno <- geno_mat_t[, win_idx, drop = FALSE]

    shrunk_cor <- lw_shrink_lw(win_geno)

    eigs <- eigen(shrunk_cor, symmetric = TRUE, only.values = TRUE)$values
    eigs[eigs < 0] <- 0

    m_eff <- m_eff + count_eigenvalues(eigs, var_thresh)
    start <- stop + 1L
  }

  return(m_eff)
}
```

# 10. Run All R Implementations

Three methods are compared against Python:

- **Method A** (`"basic"`): new `bigQTL` default — `lw_shrink_geno()` (LW 2004 on biased
  genotype covariance, no extra dependencies)
- **Method B** (`"nlshrink"`): new `bigQTL` optional — `nlshrink::linshrink_cov()`
  (requires `nlshrink`)
- **Method C** (`lw_faithful`): local validation helper using
  `cvCovEst::linearShrinkLWEst` (requires `cvCovEst`), operates on the
  in-memory genotype matrix

```{r run-r-all}
CIS_DIST   <- 1e6
VAR_THRESH <- 0.99
EIGEN_WIN  <- 200L

has_nlshrink <- requireNamespace("nlshrink",  quietly = TRUE)
has_cvcov    <- requireNamespace("cvCovEst",  quietly = TRUE)

r_results <- lapply(seq_len(nrow(features_coord)), function(i) {
  gene_row <- features_coord[i, ]

  cis_result <- get_cis_snps(
    bigsnp      = bigsnp_obj,
    pheno_chr   = gene_row$chromosome,
    pheno_start = gene_row$start,
    pheno_end   = gene_row$end,
    cis_window  = CIS_DIST
  )
  idx   <- cis_result$indices
  n_cis <- length(idx)

  ## Method A: bigQTL basic (lw_shrink_geno, no extra deps)
  m_eff_basic <- eigenMT_gene(
    bigsnp           = bigsnp_obj,
    snp_indices      = idx,
    ind.row          = ind.row,
    var_thresh       = VAR_THRESH,
    window           = EIGEN_WIN,
    shrinkage_method = "basic"
  )

  ## Method B: bigQTL nlshrink (optional)
  m_eff_nlshrink <- if (has_nlshrink) {
    eigenMT_gene(
      bigsnp           = bigsnp_obj,
      snp_indices      = idx,
      ind.row          = ind.row,
      var_thresh       = VAR_THRESH,
      window           = EIGEN_WIN,
      shrinkage_method = "nlshrink"
    )
  } else NA_integer_

  ## Method C: Python-faithful local helper (cvCovEst, optional)
  m_eff_lw <- if (has_cvcov) {
    eigenMT_gene_lw(
      geno_mat_t  = geno_mat_t,
      snp_indices = idx,
      var_thresh  = VAR_THRESH,
      window      = EIGEN_WIN
    )
  } else NA_integer_

  data.frame(
    gene           = gene_row$feature_name,
    n_cis_snps     = n_cis,
    m_eff_basic    = m_eff_basic,
    m_eff_nlshrink = m_eff_nlshrink,
    m_eff_lw       = m_eff_lw,
    stringsAsFactors = FALSE
  )
})

r_df <- do.call(rbind, r_results)
message(sprintf("R eigenMT: %d genes", nrow(r_df)))
print(head(r_df))
```

# 11. Unit Tests

```{r unit-tests}
n_failures <- 0L

## 11a. count_eigenvalues vs Python find_num_eigs ----------------------------
py_find_num_eigs <- function(eigenvalues, variance, var_thresh) {
  eigenvalues <- sort(eigenvalues, decreasing = TRUE)
  running_sum <- 0; counter <- 0L
  while (running_sum < variance * var_thresh) {
    running_sum <- running_sum + eigenvalues[counter + 1L]
    counter     <- counter + 1L
  }
  counter
}
test_eigs <- c(3.7, 2.3, 2, 1.5, 1.1, 1, 0.8, 0.5, 0.4, 0.3, 0.2, 0.1, 0.05, 0.05)
r_c  <- count_eigenvalues(test_eigs, 0.99)
py_c <- py_find_num_eigs(test_eigs, length(test_eigs), 0.99)
if (r_c != py_c) {
  n_failures <- n_failures + 1L
  message(sprintf("  FAIL count_eigenvalues: R=%d Python=%d", r_c, py_c))
} else {
  message(sprintf("  PASS count_eigenvalues (both = %d)", r_c))
}

## 11b. big_cor vs base cor() ------------------------------------------------
test_cols <- seq_len(min(30L, N_SNPS))
base_cor  <- cor(geno_mat_t[, test_cols])
fbm_cor   <- bigstatsr::big_cor(geno_fbm, ind.row = ind.row, ind.col = test_cols)[]
frob      <- sqrt(sum((base_cor - fbm_cor)^2))
if (frob >= 1e-6) {
  n_failures <- n_failures + 1L
  message(sprintf("  FAIL big_cor vs cor(): Frob=%.2e", frob))
} else {
  message(sprintf("  PASS big_cor vs cor() (Frob=%.2e)", frob))
}

## 11c. lw_shrink_geno sanity (current bigQTL "basic") ----------------------
test_win <- geno_mat_t[, seq_len(min(20L, N_SNPS)), drop = FALSE]
basic_cor <- lw_shrink_geno(test_win)
eigs_basic <- eigen(basic_cor, symmetric = TRUE, only.values = TRUE)$values
if (!all(eigs_basic >= -1e-10)) {
  n_failures <- n_failures + 1L; message("  FAIL lw_shrink_geno: not PSD")
} else if (!all(abs(diag(basic_cor) - 1) <= 1e-10)) {
  n_failures <- n_failures + 1L; message("  FAIL lw_shrink_geno: diagonal != 1")
} else {
  message("  PASS lw_shrink_geno (PSD, diagonal = 1)")
}

## 11d. lw_shrink_lw sanity (cvCovEst, optional) ----------------------------
if (has_cvcov) {
  lw_cor  <- lw_shrink_lw(test_win)
  eigs_lw <- eigen(lw_cor, symmetric = TRUE, only.values = TRUE)$values
  if (!all(eigs_lw >= -1e-10)) {
    n_failures <- n_failures + 1L; message("  FAIL lw_shrink_lw: not PSD")
  } else if (!all(abs(diag(lw_cor) - 1) <= 1e-10)) {
    n_failures <- n_failures + 1L; message("  FAIL lw_shrink_lw: diagonal != 1")
  } else {
    message("  PASS lw_shrink_lw (PSD, diagonal = 1)")
  }
} else {
  message("  SKIP lw_shrink_lw (cvCovEst not installed)")
}

## 11e. eigenMT_gene edge cases (basic method) --------------------------------
m1 <- eigenMT_gene(bigsnp_obj, snp_indices = 1L,         ind.row = ind.row)
m0 <- eigenMT_gene(bigsnp_obj, snp_indices = integer(0), ind.row = ind.row)
if (m1 != 1L) { n_failures <- n_failures + 1L; message("  FAIL single-SNP M_eff") }
if (m0 != 0L) { n_failures <- n_failures + 1L; message("  FAIL zero-SNP M_eff")   }
if (m1 == 1L && m0 == 0L) message("  PASS eigenMT_gene edge cases")

message(sprintf("\nUnit tests: %d failure(s)", n_failures))
```

# 12. Full Comparison: Python vs R Methods

```{r compare}
## Identify gene column in Python output
py_gene_col <- intersect(colnames(python_results),
                         c("gene", "geneid", "phenotype_id"))
if (length(py_gene_col) == 0L)
  stop("Cannot identify gene col in Python output: ",
       paste(colnames(python_results), collapse = ", "))
py_gene_col <- py_gene_col[1L]

comparison <- do.call(rbind, lapply(r_df$gene, function(gene) {
  r_row  <- r_df[r_df$gene == gene, ]
  py_row <- python_results[python_results[[py_gene_col]] == gene, ]
  if (nrow(r_row) == 0L || nrow(py_row) == 0L) return(NULL)
  m_py  <- py_row$TESTS[1L]
  m_b   <- r_row$m_eff_basic[1L]
  m_nls <- r_row$m_eff_nlshrink[1L]
  m_lw  <- r_row$m_eff_lw[1L]
  data.frame(
    gene              = gene,
    n_cis_snps        = r_row$n_cis_snps[1L],
    m_eff_py          = m_py,
    m_eff_basic       = m_b,
    m_eff_nlshrink    = m_nls,
    m_eff_lw          = m_lw,
    rel_diff_basic    = abs(m_b   - m_py) / max(m_py, 1L),
    rel_diff_nlshrink = if (!is.na(m_nls)) abs(m_nls - m_py) / max(m_py, 1L) else NA_real_,
    rel_diff_lw       = if (!is.na(m_lw))  abs(m_lw  - m_py) / max(m_py, 1L) else NA_real_,
    stringsAsFactors = FALSE
  )
}))

message(sprintf("Comparison: %d genes", nrow(comparison)))
print(comparison)

message(sprintf(
  "\nM_eff means — Python: %.1f | basic: %.1f | nlshrink: %s | lw (cvCovEst): %s",
  mean(comparison$m_eff_py),
  mean(comparison$m_eff_basic),
  if (!all(is.na(comparison$m_eff_nlshrink)))
    sprintf("%.1f", mean(comparison$m_eff_nlshrink, na.rm = TRUE)) else "n/a",
  if (!all(is.na(comparison$m_eff_lw)))
    sprintf("%.1f", mean(comparison$m_eff_lw, na.rm = TRUE)) else "n/a"
))

message(sprintf(
  "Relative diff vs Python — basic: median=%.0f%% max=%.0f%%",
  100 * median(comparison$rel_diff_basic, na.rm = TRUE),
  100 * max(comparison$rel_diff_basic, na.rm = TRUE)
))
if (!all(is.na(comparison$rel_diff_nlshrink)))
  message(sprintf(
    "                       nlshrink: median=%.0f%% max=%.0f%%",
    100 * median(comparison$rel_diff_nlshrink, na.rm = TRUE),
    100 * max(comparison$rel_diff_nlshrink, na.rm = TRUE)
  ))
if (!all(is.na(comparison$rel_diff_lw)))
  message(sprintf(
    "                         lw (cvCovEst): median=%.0f%% max=%.0f%%",
    100 * median(comparison$rel_diff_lw, na.rm = TRUE),
    100 * max(comparison$rel_diff_lw, na.rm = TRUE)
  ))

MEFF_TOL <- 0.20
bad_basic <- comparison[comparison$rel_diff_basic > MEFF_TOL, ]
if (nrow(bad_basic) > 0L) {
  message(sprintf("WARNING: %d gene(s) with basic M_eff diff > %d%%:",
                  nrow(bad_basic), as.integer(MEFF_TOL * 100)))
  print(bad_basic[, c("gene", "n_cis_snps", "m_eff_py", "m_eff_basic",
                       "rel_diff_basic")])
} else {
  message(sprintf("All %d genes: basic within %d%% of Python",
                  nrow(comparison), as.integer(MEFF_TOL * 100)))
}
```

# 13. Clean Up

```{r cleanup}
unlink(tmp_dir, recursive = TRUE)
message("Temp files removed.")
```
