---
title: "eigenMT Validation: R vs Python"
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Overview

This vignette validates the R implementation of eigenMT (implemented in
`bigQTL`) against the **original** Python eigenMT
([joed3/eigenMT](https://github.com/joed3/eigenMT)) by running both on the
real example dataset shipped with that repository.

The example data (`example.tar.gz`) contains a real Matrix-eQTL run on
chromosome 19: genotype matrix, SNP positions, phenotype positions, and
cis-eQTL results. We download the archive at run-time, decompress it, and
feed the same files to both implementations.

All code chunks use `eval = FALSE` and are intended to be run interactively.

## Shrinkage methods

`bigQTL::eigenMT_gene()` supports two shrinkage methods via the
`shrinkage_method` argument:

| Method | Description | Package dependency | Expected accuracy vs Python |
|--------|-------------|--------------------|-----------------------------|
| `"basic"` (default) | Analytical Ledoit-Wolf 2004 shrinkage applied to the biased sample covariance of the raw genotype matrix. Structurally identical to Python `sklearn.covariance.LedoitWolf`. | None (base R only) | Very close |
| `"nlshrink"` | Non-parametric linear shrinkage via `nlshrink::linshrink_cov()` (Ledoit & Wolf 2017). Ports the [ImmuneAxisa/eigenMT](https://github.com/ImmuneAxisa/eigenMT) R implementation. | `nlshrink` (in Suggests) | Closest to Python |

Use `shrinkage_method = "nlshrink"` when you need results that most closely
match the Python implementation. The `"basic"` method gives very close results
with no additional dependencies.

## Prerequisites

You need a conda environment named `"eigenMT"` with the required Python
packages. Create it once with:

```{r conda-setup}
library(reticulate)
conda_create(
  "eigenMT",
  packages = c("numpy", "scipy", "scikit-learn", "pandas"),
  python_version = "3.8"
)
```

To use the `"nlshrink"` method, install the nlshrink package:

```{r nlshrink-install}
install.packages("nlshrink")
```

Then load the packages:

```{r load-pkg}
library(bigQTL)
library(bigstatsr)
library(bigsnpr)
library(reticulate)
```

# 1. Download and Decompress the Example Data

The example archive is fetched from the ImmuneAxisa/eigenMT GitHub repository,
which includes outputs of the testing. We decompress it into a temporary
directory.

```{r download-example}
tmp_dir <- file.path(tempdir(), "eigenMT_validation")
dir.create(tmp_dir, recursive = TRUE, showWarnings = FALSE)

## Download the archive from ImmuneAxisa/eigenMT (includes test outputs)
tgz_url  <- paste0(
  "https://github.com/ImmuneAxisa/eigenMT/raw/",
  "0972fd84d6a6dd85d148ee8f2eb320d4a1bc027f/example.tar.gz"
)
tgz_file <- file.path(tmp_dir, "example.tar.gz")

utils::download.file(tgz_url, destfile = tgz_file, mode = "wb",
                     quiet = FALSE)
message(sprintf("Downloaded %.1f MB to %s",
                file.info(tgz_file)$size / 1e6, tgz_file))

## Decompress
utils::untar(tgz_file, exdir = tmp_dir)

## The archive unpacks to an "example/" sub-directory
example_dir <- file.path(tmp_dir, "example")
stopifnot(dir.exists(example_dir))
message("Example data files:")
print(list.files(example_dir))
```

After decompression the `example/` folder contains:

- `genotypes.txt`     — SNPs × samples genotype matrix
- `gen.positions.txt` — SNP id / chr / position
- `phe.positions.txt` — gene id / chr / start / end
- `cis.eqtls.txt`     — Matrix-eQTL output (one row per cis SNP–gene pair)

# 2. Download the Python Script

We fetch `eigenMT_fixed.py` from the ImmuneAxisa/eigenMT repository, which
contains a corrected version that works with scikit-learn ≥ 1.2.

> **Note:** The original `joed3/eigenMT` script suffers from a silent
> `np.matrix` deprecation in scikit-learn ≥ 1.2 that makes
> `sklearn.covariance.LedoitWolf.fit()` raise a `TypeError`, which is caught
> by a bare `except:` clause and falls through to a degenerate all-1s
> correlation matrix (m_eff = 1 per window). The fixed version converts
> genotype arrays correctly before fitting.

```{r download-python}
py_script_url <- paste0(
  "https://raw.githubusercontent.com/ImmuneAxisa/eigenMT/",
  "0972fd84d6a6dd85d148ee8f2eb320d4a1bc027f/eigenMT_fixed.py"
)
py_script_file <- file.path(tmp_dir, "eigenMT.py")

utils::download.file(py_script_url, destfile = py_script_file,
                     quiet = FALSE)
message("Downloaded eigenMT_fixed.py as eigenMT.py")
```

# 3. Inspect the Example Data

```{r inspect-data}
## Genotype matrix: first column is the SNP id
geno_raw <- read.table(file.path(example_dir, "genotypes.txt"),
                       header = TRUE, sep = "\t",
                       check.names = FALSE, stringsAsFactors = FALSE)

message(sprintf("Genotype matrix: %d SNPs x %d samples (+1 id col)",
                nrow(geno_raw), ncol(geno_raw) - 1L))

## SNP positions: columns  snp / chr / pos
gen_pos <- read.table(file.path(example_dir, "gen.positions.txt"),
                      header = TRUE, sep = "\t",
                      stringsAsFactors = FALSE)

message(sprintf("SNP positions: %d rows; chromosomes: %s",
                nrow(gen_pos),
                paste(unique(gen_pos[[2L]]), collapse = ", ")))

## Phenotype positions: columns  geneid / chr / s1 / s2
phe_pos <- read.table(file.path(example_dir, "phe.positions.txt"),
                      header = TRUE, sep = "\t",
                      stringsAsFactors = FALSE)
message(sprintf("Gene positions: %d genes", nrow(phe_pos)))

## cis-eQTL results (Matrix-eQTL format)
eqtl <- read.table(file.path(example_dir, "cis.eqtls.txt"),
                   header = TRUE, sep = "\t",
                   stringsAsFactors = FALSE)
message(sprintf("cis-eQTL tests: %d rows; columns: %s",
                nrow(eqtl), paste(colnames(eqtl), collapse = ", ")))
```

# 4. Run Python eigenMT

We run the fixed `eigenMT.py` via `reticulate::conda_run2()` using the
`"eigenMT"` conda environment. `--CHROM 19` matches the example data.

```{r run-python}
py_out_file <- file.path(tmp_dir, "eigenMT_output_py.txt")

reticulate::conda_run2(
  "python",
  args = c(
    py_script_file,
    "--CHROM",      "19",
    "--QTL",        file.path(example_dir, "cis.eqtls.txt"),
    "--GEN",        file.path(example_dir, "genotypes.txt"),
    "--GENPOS",     file.path(example_dir, "gen.positions.txt"),
    "--PHEPOS",     file.path(example_dir, "phe.positions.txt"),
    "--OUT",        py_out_file,
    "--cis_dist",   "1000000",
    "--var_thresh", "0.99",
    "--window",     "200"
  ),
  envname = "eigenMT"
)

python_results <- read.table(py_out_file, header = TRUE,
                             sep = "\t", stringsAsFactors = FALSE)
message(sprintf("Python eigenMT: %d genes processed", nrow(python_results)))
print(head(python_results))
```

# 5. Build a bigSNP Object from the Example Genotype Matrix

The R implementation works on FBM-backed genotype matrices (bigSNP objects).
We construct one from the same flat-file genotypes used by Python, applying
the same missing-value imputation (column mean) that Python performs.

```{r build-bigsnp}
## Sample IDs: columns 2..end of the genotype matrix
sample_ids <- colnames(geno_raw)[-1L]
N_SAMPLES  <- length(sample_ids)

## SNP ids and integer genotype matrix (SNPs × samples)
snp_ids  <- as.character(geno_raw[[1L]])
geno_mat <- as.matrix(geno_raw[, -1L, drop = FALSE])
storage.mode(geno_mat) <- "integer"

## Transpose to samples × SNPs for the FBM
geno_mat_t <- t(geno_mat)
N_SNPS     <- ncol(geno_mat_t)

## Match SNP positions from gen.positions.txt
## Column order: snp_id / chr / pos  (use positional indexing for robustness)
snp_order <- match(snp_ids, as.character(gen_pos[[1L]]))
if (any(is.na(snp_order))) {
  warning(sprintf(
    "%d SNP IDs in genotypes.txt not found in gen.positions.txt",
    sum(is.na(snp_order))
  ))
}
chr_vec <- as.character(gen_pos[[2L]][snp_order])
pos_vec <- as.numeric( gen_pos[[3L]][snp_order])

## Impute NA genotypes with per-SNP column means
## (mirrors Python: genos[genos == -1] = np.mean(genos[genos != -1]))
for (j in seq_len(N_SNPS)) {
  col_j   <- geno_mat_t[, j]
  na_mask <- is.na(col_j)
  if (any(na_mask)) {
    geno_mat_t[na_mask, j] <- as.integer(round(mean(col_j, na.rm = TRUE)))
  }
}

## Create FBM
fbm_file <- file.path(tmp_dir, "geno")
geno_fbm <- bigstatsr::FBM(
  nrow = N_SAMPLES, ncol = N_SNPS, type = "integer",
  backingfile = fbm_file,
  create_bk   = TRUE
)
geno_fbm[] <- geno_mat_t

## Assemble bigSNP-like object
snp_map <- data.frame(
  chromosome   = chr_vec,
  marker.ID    = snp_ids,
  genetic.dist = rep(0, N_SNPS),
  physical.pos = pos_vec,
  allele1      = rep("A", N_SNPS),
  allele2      = rep("C", N_SNPS),
  stringsAsFactors = FALSE
)

snp_fam <- data.frame(
  family.ID   = sample_ids,
  sample.ID   = sample_ids,
  paternal.ID = rep(0L, N_SAMPLES),
  maternal.ID = rep(0L, N_SAMPLES),
  sex         = rep(0L, N_SAMPLES),
  affection   = rep(0L, N_SAMPLES),
  stringsAsFactors = FALSE
)

bigsnp_obj <- structure(
  list(genotypes = geno_fbm, map = snp_map, fam = snp_fam),
  class = "bigSNP"
)
ind.row <- seq_len(N_SAMPLES)

message(sprintf("bigSNP object: %d samples, %d SNPs", N_SAMPLES, N_SNPS))
```

# 6. Build Gene Coordinates from phe.positions.txt

We restrict to chromosome 19 genes that appear in the cis-eQTL file,
mirroring Python's `--CHROM 19` filter.

```{r build-features}
## Second column of the eQTL file is the gene id
genes_in_eqtl <- unique(eqtl[[2L]])

## phe.positions.txt: geneid / chr / s1 / s2
features_coord <- data.frame(
  feature_name = as.character(phe_pos[[1L]]),
  chromosome   = as.character(phe_pos[[2L]]),
  start        = as.numeric(phe_pos[[3L]]),
  end          = as.numeric(phe_pos[[4L]]),
  stringsAsFactors = FALSE
)

## Keep chr19 genes that appear in the eQTL file
features_coord <- features_coord[
  features_coord$chromosome == "19" &
    features_coord$feature_name %in% genes_in_eqtl, ]

message(sprintf("Genes for R eigenMT: %d", nrow(features_coord)))
print(head(features_coord))
```

# 7. Run R eigenMT — Both Methods

We run both `"basic"` (default, no extra deps) and `"nlshrink"` (closest to
Python, requires `nlshrink` package) methods for comparison.

```{r run-r-eigenmt}
CIS_DIST   <- 1e6
VAR_THRESH <- 0.99
EIGEN_WIN  <- 200L

run_eigenmt_r <- function(method) {
  lapply(seq_len(nrow(features_coord)), function(i) {
    gene_row <- features_coord[i, ]

    cis_result <- get_cis_snps(
      bigsnp      = bigsnp_obj,
      pheno_chr   = gene_row$chromosome,
      pheno_start = gene_row$start,
      pheno_end   = gene_row$end,
      cis_window  = CIS_DIST
    )

    m_eff <- eigenMT_gene(
      bigsnp           = bigsnp_obj,
      snp_indices      = cis_result$indices,
      ind.row          = ind.row,
      var_thresh       = VAR_THRESH,
      window           = EIGEN_WIN,
      shrinkage_method = method
    )

    data.frame(
      gene       = gene_row$feature_name,
      n_cis_snps = length(cis_result$indices),
      m_eff      = m_eff,
      stringsAsFactors = FALSE
    )
  }) |> do.call(what = rbind)
}

## Basic method (default, no extra dependencies)
r_basic <- run_eigenmt_r("basic")
message(sprintf("R eigenMT (basic):    %d genes processed", nrow(r_basic)))
print(head(r_basic))

## nlshrink method (requires nlshrink package, closest to Python)
r_nlshrink <- run_eigenmt_r("nlshrink")
message(sprintf("R eigenMT (nlshrink): %d genes processed", nrow(r_nlshrink)))
print(head(r_nlshrink))
```

# 8. Unit Tests

These tests verify internal consistency of the R helpers, independent of
the Python comparison.

```{r unit-tests}
n_failures <- 0L

## 8a. count_eigenvalues vs reference Python find_num_eigs -----------------
py_find_num_eigs <- function(eigenvalues, variance, var_thresh) {
  eigenvalues <- sort(eigenvalues, decreasing = TRUE)
  running_sum <- 0
  counter     <- 0L
  while (running_sum < variance * var_thresh) {
    running_sum <- running_sum + eigenvalues[counter + 1L]
    counter     <- counter + 1L
  }
  counter
}

test_eigs <- c(3.7, 2.3, 2, 1.5, 1.1, 1, 0.8, 0.5, 0.4, 0.3, 0.2, 0.1, 0.05, 0.05)
r_count  <- count_eigenvalues(test_eigs, 0.99)
py_count <- py_find_num_eigs(test_eigs, length(test_eigs), 0.99)
if (r_count != py_count) {
  n_failures <- n_failures + 1L
  message(sprintf("  FAIL count_eigenvalues: R=%d Python=%d", r_count, py_count))
} else {
  message(sprintf("  PASS count_eigenvalues (both = %d)", r_count))
}

## 8b. lw_shrink_geno sanity checks -----------------------------------------
test_cols <- seq_len(min(20L, N_SNPS))
geno_sub  <- geno_mat_t[, test_cols, drop = FALSE]
storage.mode(geno_sub) <- "double"
shrunk <- lw_shrink_geno(geno_sub)
eigs   <- eigen(shrunk, symmetric = TRUE, only.values = TRUE)$values
if (!all(eigs >= -1e-10)) {
  n_failures <- n_failures + 1L
  message("  FAIL lw_shrink_geno: not PSD")
} else if (!all(abs(diag(shrunk) - 1) <= 1e-10)) {
  n_failures <- n_failures + 1L
  message("  FAIL lw_shrink_geno: diagonal != 1")
} else {
  message("  PASS lw_shrink_geno (PSD, diagonal = 1)")
}

## 8c. eigenMT_gene edge cases ----------------------------------------------
m1 <- eigenMT_gene(bigsnp_obj, snp_indices = 1L,         ind.row = ind.row)
m0 <- eigenMT_gene(bigsnp_obj, snp_indices = integer(0), ind.row = ind.row)
if (m1 != 1L) { n_failures <- n_failures + 1L; message("  FAIL single-SNP M_eff") }
if (m0 != 0L) { n_failures <- n_failures + 1L; message("  FAIL zero-SNP M_eff")   }
if (m1 == 1L && m0 == 0L) message("  PASS eigenMT_gene edge cases")

message(sprintf("\nUnit tests: %d failure(s)", n_failures))
```

# 9. Compare R vs Python M_eff

We join on gene ID and compare both R methods against Python.

```{r compare}
## Robustly detect the gene column name in Python output
py_gene_col <- intersect(colnames(python_results),
                         c("gene", "geneid", "phenotype_id"))
if (length(py_gene_col) == 0L) {
  stop(
    "Cannot identify gene column in Python eigenMT output. Columns: ",
    paste(colnames(python_results), collapse = ", ")
  )
}
py_gene_col <- py_gene_col[1L]

build_comparison <- function(r_df, method_label) {
  do.call(rbind, lapply(r_df$gene, function(gene) {
    r_row  <- r_df[r_df$gene == gene, ]
    py_row <- python_results[python_results[[py_gene_col]] == gene, ]
    if (nrow(r_row) == 0L || nrow(py_row) == 0L) return(NULL)
    m_r  <- r_row$m_eff[1L]
    m_py <- py_row$TESTS[1L]
    data.frame(
      gene       = gene,
      method     = method_label,
      n_cis_snps = r_row$n_cis_snps[1L],
      m_eff_r    = m_r,
      m_eff_py   = m_py,
      abs_diff   = abs(m_r - m_py),
      rel_diff   = abs(m_r - m_py) / max(m_py, 1L),
      stringsAsFactors = FALSE
    )
  }))
}

cmp_basic    <- build_comparison(r_basic,    "basic")
cmp_nlshrink <- build_comparison(r_nlshrink, "nlshrink")

print_summary <- function(cmp, label) {
  message(sprintf("\n--- Method: %s ---", label))
  message(sprintf(
    "M_eff summary — R: mean=%.1f sd=%.1f | Python: mean=%.1f sd=%.1f",
    mean(cmp$m_eff_r),  sd(cmp$m_eff_r),
    mean(cmp$m_eff_py), sd(cmp$m_eff_py)
  ))
  message(sprintf(
    "Relative diff — median=%.1f%%  max=%.1f%%",
    100 * median(cmp$rel_diff),
    100 * max(cmp$rel_diff)
  ))
}

print_summary(cmp_basic,    "basic")
print_summary(cmp_nlshrink, "nlshrink")

## Side-by-side comparison table
comparison <- merge(
  cmp_basic[, c("gene", "n_cis_snps", "m_eff_py",
                "m_eff_r", "rel_diff")],
  cmp_nlshrink[, c("gene", "m_eff_r", "rel_diff")],
  by = "gene", suffixes = c("_basic", "_nlshrink")
)
colnames(comparison)[colnames(comparison) == "m_eff_r_basic"]      <- "m_eff_basic"
colnames(comparison)[colnames(comparison) == "m_eff_r_nlshrink"]   <- "m_eff_nlshrink"
colnames(comparison)[colnames(comparison) == "rel_diff_basic"]      <- "rel_diff_basic"
colnames(comparison)[colnames(comparison) == "rel_diff_nlshrink"]   <- "rel_diff_nlshrink"

message(sprintf("\nComparison table (%d genes):", nrow(comparison)))
print(comparison)
```

# 10. Clean Up

```{r cleanup}
unlink(tmp_dir, recursive = TRUE)
message("Temp files removed.")
```
